---
title: "Admixture"
output: html_document
date: "2026-05-07"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load libraries
```{r}
library(tidyverse)
library(ggthemes)
library(ggrepel)
library(patchwork)
library(lfmm)
library(ggplot2)
library(dplyr)
```
Read historical data
```{r}
his_data = read.delim("data/Historical_climate_caladapt_1950_2004.tsv", header = TRUE)
```

Read Future Climate data
```{r}
fut_data = read.delim("data/Future_climate_ppt_caladapt_2040_2070.tsv", header = TRUE)
```

```{r}
# We would be using this new data and not the old one.
geno_23 = read.delim("data/snp_allele_freq_mat.csv", header = TRUE, sep = ",")
pops =colnames(geno_23)[-1]
npop = length(pops)
```


#Function for LFMM
```{r}
run_lasso <- function(X=NULL,Y=NULL,K=5) {
   mod.lfmm_lasso <- lfmm_lasso(Y = Y, 
                        X = X, 
                        K = K,
                        nozero.prop = 0.01)
   pv_lasso <- lfmm_test(Y = Y, 
                 X = X, 
                 lfmm = mod.lfmm_lasso, 
                 calibrate = "gif")
  
  # Made changes here. It now returns both the objects
   return(list(pv = pv_lasso, mod = mod.lfmm_lasso))
}
```


#Function for qqplot
```{r}
plot_qq <- function(pvalue_data = NULL,
                    index = 1){
  pvalues.raw = pvalue_data$pvalue[,index]
  
n = length(sort(pvalues.raw))
df = tibble(observed = sort(-log10(pvalues.raw)),
expected = sort(-log10(ppoints(n))),
clower   = -log10(qbeta(p = (1 - 0.95) / 2, shape1 = 1:n, shape2 = n:1)),
cupper   = -log10(qbeta(p = (1 + 0.95) / 2, shape1 = 1:n, shape2 = n:1))
)

# Convert p-values to 1-df chi-squared statistics
chi2 <- qchisq(pvalues.raw, df=1, lower.tail=FALSE)
# Calculate lambda
lambda <- median(chi2) / qchisq(0.5, df=1)
print(lambda)

y_max_print = df |>
  slice_max(observed, n=1) |>
  pull(observed)
  lambda = round(lambda, digits =2)


p = df |>
ggplot() +
geom_ribbon(aes(x = expected, ymin = clower, ymax = cupper), alpha = 0.1) +
geom_point(aes(expected, observed), shape = 1, size = 3) +
geom_abline(intercept = 0, slope = 1, alpha = 0.5) +
theme_bw() +
annotate(geom = "text", x=1, y = y_max_print-0.5, label = paste0("Lambda = ", lambda))+
labs(x = expression(paste("Expected -log"[10], plain(P))),
y = expression(paste("Observed -log"[10], plain(P))))

return(p)

}
```

#Function for genomic offset
```{r}
calculate_genomic_offset = function(allele_freq, 
                                     lfmm_effects, 
                                     env_current, 
                                     env_future,
                                     snp_filter = NULL,
                                     method = "euclidean") {
  
  # Dimensions check
  n_pop = nrow(allele_freq)
  n_snp = ncol(allele_freq)
  n_env = ncol(env_current)
  
  cat("\n=== Genomic Offset Calculation ===\n")
  cat("Populations:", n_pop, "\n")
  cat("SNPs (total):", n_snp, "\n")
  cat("Environmental variables:", n_env, "\n")
  
  # Apply SNP filter if provided
  if (!is.null(snp_filter)) {
    allele_freq = allele_freq[, snp_filter, drop = FALSE]
    lfmm_effects = lfmm_effects[snp_filter, , drop = FALSE]
    cat("SNPs (after filtering):", sum(snp_filter), "\n")
  }
  
  # Ensure environmental variables are standardized
  # Use same scaling for current and future based on current
  env_means = colMeans(env_current)
  env_sds = apply(env_current, 2, sd)
  
  env_current_scaled = scale(env_current, center = env_means, scale = env_sds)
  env_future_scaled = scale(env_future, center = env_means, scale = env_sds)
  
  cat("\nEnvironmental variables standardized using current climate statistics\n")
  
  # ========================================================================
  # STEP 1: Calculate genetic values under current climate
  # ========================================================================
  # For each population, calculate adaptive genetic value
  # genetic_value = sum over SNPs and env vars of:
  #   (allele_freq × effect_size × env_value)
  
  genetic_values_current = matrix(0, nrow = n_pop, ncol = n_env)
  
  for (i in 1:n_pop) {
    for (j in 1:n_env) {
      # Allele frequency × effect size for this env variable
      weighted_freqs = allele_freq[i, ] * lfmm_effects[, j]
      # Weighted by environmental value
      genetic_values_current[i, j] = sum(weighted_freqs) * env_current_scaled[i, j]
    }
  }
  
  # Total genetic value across all environmental axes
  genetic_value_current_total = rowSums(genetic_values_current)
  
  cat("\nCurrent genetic values calculated\n")
  
  # ========================================================================
  # STEP 2: Calculate genetic values under future climate
  # ========================================================================
  
  genetic_values_future = matrix(0, nrow = n_pop, ncol = n_env)
  
  for (i in 1:n_pop) {
    for (j in 1:n_env) {
      # Same allele frequencies (no evolution assumed)
      # But different environmental values
      weighted_freqs = allele_freq[i, ] * lfmm_effects[, j]
      genetic_values_future[i, j] = sum(weighted_freqs) * env_future_scaled[i, j]
    }
  }
  
  genetic_value_future_total = rowSums(genetic_values_future)
  
  cat("Future genetic values calculated\n")
  
  # ========================================================================
  # STEP 3: Calculate offset
  # ========================================================================
  
  if (method == "euclidean") {
    # Euclidean distance across all environmental dimensions
    offset = sqrt(rowSums((genetic_values_future - genetic_values_current)^2))
    cat("\nOffset method: Euclidean distance\n")
    
  } else if (method == "absolute") {
    # Simple absolute difference
    offset = abs(genetic_value_future_total - genetic_value_current_total)
    cat("\nOffset method: Absolute difference\n")
    
  } else if (method == "relative") {
    # Relative change
    offset = abs((genetic_value_future_total - genetic_value_current_total) / 
                    (abs(genetic_value_current_total) + 1e-10))
    cat("\nOffset method: Relative change\n")
  }
  
  # ========================================================================
  # STEP 4: Per-variable contributions
  # ========================================================================
  
  # Calculate offset contribution from each environmental variable
  offset_by_var = sqrt((genetic_values_future - genetic_values_current)^2)
  colnames(offset_by_var) = colnames(env_current)
  
  cat("\nOffset calculation complete!\n")
  cat("Mean offset:", round(mean(offset), 4), "\n")
  cat("Range:", round(min(offset), 4), "-", round(max(offset), 4), "\n")
  
  # Return results
  return(list(
    offset = offset,
    offset_by_variable = offset_by_var,
    genetic_values_current = genetic_value_current_total,
    genetic_values_future = genetic_value_future_total,
    genetic_values_current_by_var = genetic_values_current,
    genetic_values_future_by_var = genetic_values_future,
    env_change = env_future_scaled - env_current_scaled,
    n_snps_used = ncol(allele_freq)
  ))
}
```













#New Frequency Matrix with alpha = 0.1
```{r}
alpha <- 0.1

new_LV1 = ((1 - alpha) * geno_23$LV1) + (alpha * geno_23$SQ1)
new_allele_mat <- geno_23
new_allele_mat$LV1 = new_LV1
```

```{r}
plot(new_allele_mat$LV1, geno_23$LV1, xlab = "new LV1", ylab = "old LV1", main = "Admixture Plot")
```

#admixture plot visualizes how much the allele frequencies shift from original lv1 as alpha increases







#LFMM (leaving out LV1 from original matrix)
```{r}
Y_2023_noLV1 = as.matrix(geno_23[,-1])
Y_2023_noLV1 = t(Y_2023_noLV1)
Y_2023_noLV1 <- Y_2023_noLV1[rownames(Y_2023_noLV1) != "LV1", ]
```

```{r}
Y_2023 = as.matrix(geno_23[,-1])
Y_2023 = t(Y_2023)
```

```{r}
gcm_mean_his <- his_data %>% 
  filter(dt >= "1990-12-31") %>% 
  group_by(id, cvar, dt) %>%
  summarize(mean_val = mean(val)) %>% 
  ungroup() %>% 
  group_by(id, cvar) %>% 
  summarise(mean_val_ = mean(mean_val),
            cv = sd(mean_val)/mean(mean_val)) %>% 
  ungroup()
```


```{r}
gcm_mean = gcm_mean_his %>% 
  pivot_longer(cols = mean_val_:cv) %>% 
  unite(name, cvar:name) %>% 
  pivot_wider(names_from = name, values_from = value)
```




```{r}
X_2023_gcm_mean = gcm_mean %>% 
  ungroup() %>% 
  select(-id) %>% 
  as.matrix() %>% 
  scale()
rownames(X_2023_gcm_mean) <- gcm_mean$id
```


```{r}
common_noLV1 <- intersect(rownames(X_2023_gcm_mean), rownames(Y_2023_noLV1))
Y_2023_noLV1 <- Y_2023_noLV1[common_noLV1, ]
X_2023_noLV1 <- X_2023_gcm_mean[common_noLV1, , drop = FALSE]
```

```{r}
pv_lasso_noLV1 <- run_lasso(X_2023_noLV1, Y_2023_noLV1)
```

```{r}
p65 = plot_qq(pv_lasso_noLV1$pv, index = 1) + ggtitle("PV Lasso GCM Mean w/o LV1 - Mean Precipitation")
p66 = plot_qq(pv_lasso_noLV1$pv, index = 2) + ggtitle("PV Lasso GCM Mean w/o LV1 - CV Precipitation")
p67 = plot_qq(pv_lasso_noLV1$pv, index = 3) + ggtitle("PV Lasso GCM Mean w/o LV1 - Mean Maximum Temperature")
p68 = plot_qq(pv_lasso_noLV1$pv, index = 4) + ggtitle("PV Lasso GCM Mean w/o LV1 - CV Maximum Temperature")
p69 = plot_qq(pv_lasso_noLV1$pv, index = 5) + ggtitle("PV Lasso GCM Mean w/o LV1 - Mean Minimum Temperature")
p70 = plot_qq(pv_lasso_noLV1$pv, index = 6) + ggtitle("PV Lasso GCM Mean w/o LV1 - CV Minimum Temperature")
print(p65)
print(p66)
print(p67)
print(p68)
print(p69)
print(p70)
```
```{r}
ggsave("../Figures/pv_lasso_gcm_mean_pr_noLV1.png", plot = p65, width = 10, height = 6)
ggsave("../Figures/pv_lasso_gcm_cv_pr_noLV1.png", plot = p66, width = 10, height = 6)
ggsave("../Figures/pv_lasso_gcm_mean_tasmax_noLV1.png", plot = p67, width = 10, height = 6)
ggsave("../Figures/pv_lasso_gcm_cv_tasmax_noLV1.png", plot = p68, width = 10, height = 6)
ggsave("../Figures/pv_lasso_gcm_mean_tasmin_noLV1.png", plot = p69, width = 10, height = 6)
ggsave("../Figures/pv_lasso_gcm_cv_tasmin_noLV1.png", plot = p70, width = 10, height = 6)
```

#lfmm used to identify snps associated with 6 environmental variableswhich produces B matrix (snps x environmental variables) capture how strongly each snp responds to each climate variable
#k=5 represents the number of latent factors which account for population structure so we're not confounding true associations with population structure
















#Future mean Climate Data
```{r}
gcm_mean_fut <- fut_data %>% 
  filter(scenario == "rcp85") %>%
  group_by(id, cvar, dt) %>% 
  summarise(mean_val = mean(val)) %>% 
  ungroup() %>% 
  group_by(id, cvar) %>%
  summarise(mean_val_ = mean(mean_val),
            cv = sd(mean_val)/mean(mean_val)) %>% 
  ungroup() %>% 
  pivot_longer(cols = mean_val_:cv) %>% 
  unite(name, cvar:name) %>% 
  pivot_wider(names_from = name, values_from = value)
```


```{r}
X_gcm_fut = gcm_mean_fut %>% 
  ungroup() %>% 
  select(-id) %>% 
  as.matrix() %>% 
  scale()
```

```{r}
rownames(X_gcm_fut) = gcm_mean_fut$id
common=intersect(rownames(X_gcm_fut), rownames(Y_2023))
X_gcm_fut = X_gcm_fut[common, , drop = FALSE]
```






#genetic offset measures how mismatched a population's current genetic composition will be to future climate conditions. higher offset =  greater risk of maladaptation under climate change


#make two genetic offset plots with new and old matrices

```{r}
#Original offset
# Extract B here
B = pv_lasso_noLV1$mod$B
allele_freq = Y_2023
pop_names = rownames(allele_freq)
common = intersect(pop_names, rownames(X_2023_gcm_mean))
env_current = as.matrix(X_2023_gcm_mean[pop_names ,])
#rownames(env_current) = rownames(X_2023_gcm_mean)
env_future = as.matrix(X_gcm_fut)
#rownames(env_future) = rownames(X_gcm_fut)
lfmm_effects = B
# lfmm_pvalues = pv_2023$pvalue



offset_all_gcm_mean = calculate_genomic_offset(
  allele_freq = allele_freq,
  lfmm_effects = lfmm_effects,
  env_current = env_current,
  env_future = env_future,
  snp_filter = NULL,  # Use all SNPs
  method = "euclidean"
)
```



```{r}
results_df <- data.frame(
  population = pop_names,
  # Offsets
  offset_all_snps = offset_all_gcm_mean$offset,
  # Genetic values
  genetic_value_current_all = offset_all_gcm_mean$genetic_values_current,
  genetic_value_future_all = offset_all_gcm_mean$genetic_values_future,
  # Add environmental change magnitude
  env_change_magnitude = sqrt(rowSums((env_future - env_current)^2))
)
```

```{r}
p71 <- ggplot(results_df, aes(x = env_change_magnitude, y = offset_all_snps, label = population)) +
  geom_point(size = 3, alpha = 0.7, color = "darkred") +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  geom_text_repel()+
  theme_minimal() +
  theme(
    axis.text = element_text(size = 11),
    axis.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    x = "Environmental Change Magnitude",
    y = "Genomic Offset (ALL SNPs)",
    title = "Offset vs Environmental Change - Mean of All Models w/0 LV1 (orignal matrix)"
  )
print(p71)
```
```{r}
ggsave("../Figures/offset_gcm_mean_noLV1.png", plot = p71, width = 10, height = 6)
```




#admixed plot with alpha = 0.1
```{r}
#admixed offset
# Extract B here
B = pv_lasso_noLV1$mod$B
adx_allele_freq = as.matrix(new_allele_mat[, -1]) #remove snpid column
adx_allele_freq = t(adx_allele_freq)

# Align to common populations with env data
common_adx = intersect(
  intersect(rownames(X_2023_gcm_mean), rownames(X_gcm_fut)),
  rownames(adx_allele_freq)
)

adx_allele_freq = adx_allele_freq[common_adx, , drop = FALSE]
env_current = as.matrix(X_2023_gcm_mean[common_adx, , drop = FALSE])
#rownames(env_current) = rownames(X_2023_gcm_mean)
env_future = as.matrix(X_gcm_fut[common_adx, , drop = FALSE])
#rownames(env_future) = rownames(X_gcm_fut)
lfmm_effects = B
# lfmm_pvalues = pv_2023$pvalue




adx_offset_all_gcm_mean = calculate_genomic_offset(
  allele_freq = adx_allele_freq,
  lfmm_effects = B,
  env_current = env_current,
  env_future = env_future,
  snp_filter = NULL,  # Use all SNPs
  method = "euclidean"
)
```

```{r}
results_df_adx <- data.frame(
  population = pop_names,
  # Offsets
  offset_all_snps = adx_offset_all_gcm_mean$offset,
  # Genetic values
  genetic_value_current_all = adx_offset_all_gcm_mean$genetic_values_current,
  genetic_value_future_all = adx_offset_all_gcm_mean$genetic_values_future,
  # Add environmental change magnitude
  env_change_magnitude = sqrt(rowSums((env_future - env_current)^2))
)
```

```{r}
p72 <- ggplot(results_df_adx, aes(x = env_change_magnitude, y = offset_all_snps, label = population)) +
  geom_point(size = 3, alpha = 0.7, color = "darkred") +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  geom_text_repel()+
  theme_minimal() +
  theme(
    axis.text = element_text(size = 11),
    axis.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    x = "Environmental Change Magnitude",
    y = "Genomic Offset (ALL SNPs)",
    title = "Offset vs Environmental Change - Mean of All Models Admixed Mixture alpha = 0.1"
  )
print(p72)
```

```{r}
ggsave("../Figures/adx_offset_gcm_mean_noLV1_01.png", plot = p72, width = 10, height = 6)
```

#simulated assisted gene flow where alpha as 0.1 = lv1's allele frequencies reflect 10% contribution from sq1 and we tested upto 0.3
#trying to see if the offset is actually decreasing as we increase gene flow
#reduction in offset would mean that the admixed population's allele freqeuencies are closer to what would be selectively favored under future climate; higher chance of adapting successfully without going locally extinct


#New Allele Frequency matrix with alpha = 0.2
```{r}
alpha <- 0.2

new_LV1_02 = ((1 - alpha) * geno_23$LV1) + (alpha * geno_23$SQ1)
new_allele_mat_02 <- geno_23
new_allele_mat_02$LV1 = new_LV1_02
```

```{r}
plot(new_allele_mat_02$LV1, geno_23$LV1, xlab = "new LV1", ylab = "old LV1", main = "Admixture Plot alpha = 0.2")
```

#offset calculations
```{r}
#admixed offset
# Extract B here
B = pv_lasso_noLV1$mod$B
adx_allele_freq_02 = as.matrix(new_allele_mat_02[, -1]) #remove snpid column
adx_allele_freq_02 = t(adx_allele_freq_02)

# Align to common populations with env data
common_adx_02 = intersect(
  intersect(rownames(X_2023_gcm_mean), rownames(X_gcm_fut)),
  rownames(adx_allele_freq_02)
)

adx_allele_freq_02 = adx_allele_freq_02[common_adx_02, , drop = FALSE]
env_current_02 = as.matrix(X_2023_gcm_mean[common_adx_02, , drop = FALSE])
#rownames(env_current) = rownames(X_2023_gcm_mean)
env_future_02 = as.matrix(X_gcm_fut[common_adx_02, , drop = FALSE])
#rownames(env_future) = rownames(X_gcm_fut)
#lfmm_effects = B
# lfmm_pvalues = pv_2023$pvalue




adx_offset_all_gcm_mean_02 = calculate_genomic_offset(
  allele_freq = adx_allele_freq_02,
  lfmm_effects = B,
  env_current = env_current_02,
  env_future = env_future_02,
  snp_filter = NULL,  # Use all SNPs
  method = "euclidean"
)
```

```{r}
results_df_adx_02 <- data.frame(
  population = pop_names,
  # Offsets
  offset_all_snps = adx_offset_all_gcm_mean_02$offset,
  # Genetic values
  genetic_value_current_all = adx_offset_all_gcm_mean_02$genetic_values_current,
  genetic_value_future_all = adx_offset_all_gcm_mean_02$genetic_values_future,
  # Add environmental change magnitude
  env_change_magnitude = sqrt(rowSums((env_future - env_current)^2))
)
```

```{r}
p76 <- ggplot(results_df_adx_02, aes(x = env_change_magnitude, y = offset_all_snps, label = population)) +
  geom_point(size = 3, alpha = 0.7, color = "darkred") +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  geom_text_repel()+
  theme_minimal() +
  theme(
    axis.text = element_text(size = 11),
    axis.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    x = "Environmental Change Magnitude",
    y = "Genomic Offset (ALL SNPs)",
    title = "Offset vs Environmental Change - Mean of All Models Admixed Mixture alpha = 0.2"
  )
print(p76)
```

```{r}
ggsave("../Figures/adx_offset_gcm_mean_noLV1_02.png", plot = p76, width = 10, height = 6)
```

#New Allele Frequency matrix with alpha = 0.3
```{r}
alpha <- 0.3

new_LV1_03 = ((1 - alpha) * geno_23$LV1) + (alpha * geno_23$SQ1)
new_allele_mat_03 <- geno_23
new_allele_mat_03$LV1 = new_LV1_03
```

```{r}
plot(new_allele_mat_03$LV1, geno_23$LV1, xlab = "new LV1", ylab = "old LV1", main = "Admixture Plot alpha = 0.3")
```

#offset calculations
```{r}
#admixed offset
# Extract B here
B = pv_lasso_noLV1$mod$B
adx_allele_freq_03 = as.matrix(new_allele_mat_03[, -1]) #remove snpid column
adx_allele_freq_03 = t(adx_allele_freq_03)

# Align to common populations with env data
common_adx_03 = intersect(
  intersect(rownames(X_2023_gcm_mean), rownames(X_gcm_fut)),
  rownames(adx_allele_freq_03)
)

adx_allele_freq_03 = adx_allele_freq_03[common_adx_03, , drop = FALSE]
env_current_03 = as.matrix(X_2023_gcm_mean[common_adx_03, , drop = FALSE])
#rownames(env_current) = rownames(X_2023_gcm_mean)
env_future_03 = as.matrix(X_gcm_fut[common_adx_03, , drop = FALSE])
#rownames(env_future) = rownames(X_gcm_fut)
lfmm_effects = B
# lfmm_pvalues = pv_2023$pvalue




adx_offset_all_gcm_mean_03 = calculate_genomic_offset(
  allele_freq = adx_allele_freq_03,
  lfmm_effects = B,
  env_current = env_current_03,
  env_future = env_future_03,
  snp_filter = NULL,  # Use all SNPs
  method = "euclidean"
)
```

```{r}
results_df_adx_03 <- data.frame(
  population = pop_names,
  # Offsets
  offset_all_snps = adx_offset_all_gcm_mean_03$offset,
  # Genetic values
  genetic_value_current_all = adx_offset_all_gcm_mean_03$genetic_values_current,
  genetic_value_future_all = adx_offset_all_gcm_mean_03$genetic_values_future,
  # Add environmental change magnitude
  env_change_magnitude = sqrt(rowSums((env_future - env_current)^2))
)
```

```{r}
p77 <- ggplot(results_df_adx_03, aes(x = env_change_magnitude, y = offset_all_snps, label = population)) +
  geom_point(size = 3, alpha = 0.7, color = "darkred") +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  geom_text_repel()+
  theme_minimal() +
  theme(
    axis.text = element_text(size = 11),
    axis.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    x = "Environmental Change Magnitude",
    y = "Genomic Offset (ALL SNPs)",
    title = "Offset vs Environmental Change - Mean of All Models Admixed Mixture alpha = 0.3"
  )
print(p77)
```

```{r}
ggsave("../Figures/adx_offset_gcm_mean_noLV1_03.png", plot = p77, width = 10, height = 6)
```













