---
title: "LOO_CV"
output: html_document
date: "2026-04-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load libraries
```{r}
library(tidyverse)
library(ggthemes)
library(ggrepel)
library(patchwork)
library(lfmm)
library(ggplot2)
library(dplyr)
source("manhattan_plot.R")
```

```{r}
# We would be using this new data and not the old one.
geno_23 = read.delim("data/snp_allele_freq_mat.csv", header = TRUE, sep = ",")
pops =colnames(geno_23)[-1]
npop = length(pops)
```

Read historical data
```{r}
his_data = read.delim("data/Historical_climate_caladapt_1950_2004.tsv", header = TRUE)
```

Read Future Climate data
```{r}
fut_data = read.delim("data/Future_climate_ppt_caladapt_2040_2070.tsv", header = TRUE)
```



Read LOO_CV RDS file
```{r}
loo_cv <- readRDS("data/loo_cv.rds")
```

```{r}
comparison_table = loo_cv$comparison
```

ggplot
```{r}
comparison_plot <- 
  ggplot(comparison_table, aes(x = offset_full, y = offset_loo, label = population)) +
  geom_point(aes(color = leverage_flag)) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_text_repel() +
  labs(
    x = "Offset Full",
    y = "Offset LOO",
    title = "Offset Full vs Offset LOO"
  ) +
  theme_minimal()

comparison_plot
```
certain populations have high leverage which includes the FR, SC, IH, WL2. even with the modest environmental change, show high offset values. remove these populations during initial lfmm run and project it again for the offset calculation: 

B matrix: weighted allele frequency matrix(strength of the association between climate and snps) to capture the climate-allele relationship and project future climate data which never saw the left out population. predict the offset for the left out populations

runn lfmm on the orginal allele frequency matrix exclude lassen populations

do admixture  = new allele frequency matrix with lassen populations which will account for the mixed ancestry and use that for the weighted allele frequency matrix which is the B matrix and then project it to the future climate to get the predicted offset for the left out populations.

```{r}
run_lasso <- function(X=NULL,Y=NULL,K=5) {
   mod.lfmm_lasso <- lfmm_lasso(Y = Y, 
                        X = X, 
                        K = K,
                        nozero.prop = 0.01)
   pv_lasso <- lfmm_test(Y = Y, 
                 X = X, 
                 lfmm = mod.lfmm_lasso, 
                 calibrate = "gif")
  
  # Made changes here. It now returns both the objects
   return(list(pv = pv_lasso, mod = mod.lfmm_lasso))
}
```

Function for qqplot
```{r}
plot_qq <- function(pvalue_data = NULL,
                    index = 1){
  pvalues.raw = pvalue_data$pvalue[,index]
  
n = length(sort(pvalues.raw))
df = tibble(observed = sort(-log10(pvalues.raw)),
expected = sort(-log10(ppoints(n))),
clower   = -log10(qbeta(p = (1 - 0.95) / 2, shape1 = 1:n, shape2 = n:1)),
cupper   = -log10(qbeta(p = (1 + 0.95) / 2, shape1 = 1:n, shape2 = n:1))
)

# Convert p-values to 1-df chi-squared statistics
chi2 <- qchisq(pvalues.raw, df=1, lower.tail=FALSE)
# Calculate lambda
lambda <- median(chi2) / qchisq(0.5, df=1)
print(lambda)

y_max_print = df |>
  slice_max(observed, n=1) |>
  pull(observed)
  lambda = round(lambda, digits =2)


p = df |>
ggplot() +
geom_ribbon(aes(x = expected, ymin = clower, ymax = cupper), alpha = 0.1) +
geom_point(aes(expected, observed), shape = 1, size = 3) +
geom_abline(intercept = 0, slope = 1, alpha = 0.5) +
theme_bw() +
annotate(geom = "text", x=1, y = y_max_print-0.5, label = paste0("Lambda = ", lambda))+
labs(x = expression(paste("Expected -log"[10], plain(P))),
y = expression(paste("Observed -log"[10], plain(P))))

return(p)

}
```



Function for genomic offset
```{r}
calculate_genomic_offset = function(allele_freq, 
                                     lfmm_effects, 
                                     env_current, 
                                     env_future,
                                     snp_filter = NULL,
                                     method = "euclidean") {
  
  # Dimensions check
  n_pop = nrow(allele_freq)
  n_snp = ncol(allele_freq)
  n_env = ncol(env_current)
  
  cat("\n=== Genomic Offset Calculation ===\n")
  cat("Populations:", n_pop, "\n")
  cat("SNPs (total):", n_snp, "\n")
  cat("Environmental variables:", n_env, "\n")
  
  # Apply SNP filter if provided
  if (!is.null(snp_filter)) {
    allele_freq = allele_freq[, snp_filter, drop = FALSE]
    lfmm_effects = lfmm_effects[snp_filter, , drop = FALSE]
    cat("SNPs (after filtering):", sum(snp_filter), "\n")
  }
  
  # Ensure environmental variables are standardized
  # Use same scaling for current and future based on current
  env_means = colMeans(env_current)
  env_sds = apply(env_current, 2, sd)
  
  env_current_scaled = scale(env_current, center = env_means, scale = env_sds)
  env_future_scaled = scale(env_future, center = env_means, scale = env_sds)
  
  cat("\nEnvironmental variables standardized using current climate statistics\n")
  
  # ========================================================================
  # STEP 1: Calculate genetic values under current climate
  # ========================================================================
  # For each population, calculate adaptive genetic value
  # genetic_value = sum over SNPs and env vars of:
  #   (allele_freq × effect_size × env_value)
  
  genetic_values_current = matrix(0, nrow = n_pop, ncol = n_env)
  
  for (i in 1:n_pop) {
    for (j in 1:n_env) {
      # Allele frequency × effect size for this env variable
      weighted_freqs = allele_freq[i, ] * lfmm_effects[, j]
      # Weighted by environmental value
      genetic_values_current[i, j] = sum(weighted_freqs) * env_current_scaled[i, j]
    }
  }
  
  # Total genetic value across all environmental axes
  genetic_value_current_total = rowSums(genetic_values_current)
  
  cat("\nCurrent genetic values calculated\n")
  
  # ========================================================================
  # STEP 2: Calculate genetic values under future climate
  # ========================================================================
  
  genetic_values_future = matrix(0, nrow = n_pop, ncol = n_env)
  
  for (i in 1:n_pop) {
    for (j in 1:n_env) {
      # Same allele frequencies (no evolution assumed)
      # But different environmental values
      weighted_freqs = allele_freq[i, ] * lfmm_effects[, j]
      genetic_values_future[i, j] = sum(weighted_freqs) * env_future_scaled[i, j]
    }
  }
  
  genetic_value_future_total = rowSums(genetic_values_future)
  
  cat("Future genetic values calculated\n")
  
  # ========================================================================
  # STEP 3: Calculate offset
  # ========================================================================
  
  if (method == "euclidean") {
    # Euclidean distance across all environmental dimensions
    offset = sqrt(rowSums((genetic_values_future - genetic_values_current)^2))
    cat("\nOffset method: Euclidean distance\n")
    
  } else if (method == "absolute") {
    # Simple absolute difference
    offset = abs(genetic_value_future_total - genetic_value_current_total)
    cat("\nOffset method: Absolute difference\n")
    
  } else if (method == "relative") {
    # Relative change
    offset = abs((genetic_value_future_total - genetic_value_current_total) / 
                    (abs(genetic_value_current_total) + 1e-10))
    cat("\nOffset method: Relative change\n")
  }
  
  # ========================================================================
  # STEP 4: Per-variable contributions
  # ========================================================================
  
  # Calculate offset contribution from each environmental variable
  offset_by_var = sqrt((genetic_values_future - genetic_values_current)^2)
  colnames(offset_by_var) = colnames(env_current)
  
  cat("\nOffset calculation complete!\n")
  cat("Mean offset:", round(mean(offset), 4), "\n")
  cat("Range:", round(min(offset), 4), "-", round(max(offset), 4), "\n")
  
  # Return results
  return(list(
    offset = offset,
    offset_by_variable = offset_by_var,
    genetic_values_current = genetic_value_current_total,
    genetic_values_future = genetic_value_future_total,
    genetic_values_current_by_var = genetic_values_current,
    genetic_values_future_by_var = genetic_values_future,
    env_change = env_future_scaled - env_current_scaled,
    n_snps_used = ncol(allele_freq)
  ))
}
```


```{r}
gcm_mean_his <- his_data %>% 
  filter(dt >= "1990-12-31") %>% 
  group_by(id, cvar, dt) %>%
  summarize(mean_val = mean(val)) %>% 
  ungroup() %>% 
  group_by(id, cvar) %>% 
  summarise(mean_val_ = mean(mean_val),
            cv = sd(mean_val)/mean(mean_val)) %>% 
  ungroup()
```


```{r}
gcm_mean = gcm_mean_his %>% 
  pivot_longer(cols = mean_val_:cv) %>% 
  unite(name, cvar:name) %>% 
  pivot_wider(names_from = name, values_from = value)
```

remove the four populations
```{r}
gcm_mean = gcm_mean %>% 
  filter(!id %in% c("WL2", "IH", "FR", "SC"))
```


do this everytime running a model
```{r}
X_2023_gcm_mean = gcm_mean %>% 
  ungroup() %>% 
  select(-id) %>% 
  as.matrix() %>% 
  scale()
```

Aligning genotype and environment data to match the same populations (rows)
X_2023 = environmental variable matrix, Y_2023 = genotype matrix
```{r}
rownames(X_2023_gcm_mean) = gcm_mean$id
common = intersect(rownames(X_2023_gcm_mean), rownames(Y_2023))
Y_2023 = Y_2023[common,]
X_2023_gcm_mean = X_2023_gcm_mean[common, , drop = FALSE]
```

```{r}
pv_lasso_gcm_mean = run_lasso(X_2023_gcm_mean, Y_2023)
```




```{r}
plot_qq(pv_lasso_gcm_mean$pv, index = 1)
plot_qq(pv_lasso_gcm_mean$pv, index = 2)
plot_qq(pv_lasso_gcm_mean$pv, index = 3)
plot_qq(pv_lasso_gcm_mean$pv, index = 4)
plot_qq(pv_lasso_gcm_mean$pv, index = 5)
plot_qq(pv_lasso_gcm_mean$pv, index = 6)
```

```{r}
remove_pops = c("WL2", "IH", "FR", "SC")
```


Future mean Climate Data
```{r}
gcm_mean_fut <- fut_data %>% 
  filter(scenario == "rcp85") %>%
  group_by(id, cvar, dt) %>% 
  summarise(mean_val = mean(val)) %>% 
  ungroup() %>% 
  group_by(id, cvar) %>%
  summarise(mean_val_ = mean(mean_val),
            cv = sd(mean_val)/mean(mean_val)) %>% 
  ungroup() %>% 
  pivot_longer(cols = mean_val_:cv) %>% 
  unite(name, cvar:name) %>% 
  pivot_wider(names_from = name, values_from = value) %>% 
  filter(!id %in% remove_pops)
```


```{r}
X_gcm_fut = gcm_mean_fut %>% 
  ungroup() %>% 
  select(-id) %>% 
  as.matrix() %>% 
  scale()
```


```{r}
rownames(X_gcm_fut) = gcm_mean_fut$id
X_gcm_fut = X_gcm_fut[common, , drop = FALSE]
```



```{r}
# Extract B here
B = pv_lasso_gcm_mean$mod$B
allele_freq = Y_2023
env_current = as.matrix(X_2023_gcm_mean)
#rownames(env_current) = rownames(X_2023_gcm_mean)
env_future = as.matrix(X_gcm_fut)
#rownames(env_future) = rownames(X_gcm_fut)
lfmm_effects = B
# lfmm_pvalues = pv_2023$pvalue
pop_names = rownames(allele_freq)


offset_all_gcm_mean = calculate_genomic_offset(
  allele_freq = allele_freq,
  lfmm_effects = lfmm_effects,
  env_current = env_current,
  env_future = env_future,
  snp_filter = NULL,  # Use all SNPs
  method = "euclidean"
)
```


```{r}
results_df <- data.frame(
  population = pop_names,
  # Offsets
  offset_all_snps = offset_all_gcm_mean$offset,
  # Genetic values
  genetic_value_current_all = offset_all_gcm_mean$genetic_values_current,
  genetic_value_future_all = offset_all_gcm_mean$genetic_values_future,
  # Add environmental change magnitude
  env_change_magnitude = sqrt(rowSums((env_future - env_current)^2))
)
```

```{r}
p1 <- ggplot(results_df, aes(x = env_change_magnitude, y = offset_all_snps, label = population)) +
  geom_point(size = 3, alpha = 0.7, color = "darkred") +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  geom_text_repel()+
  theme_minimal() +
  theme(
    axis.text = element_text(size = 11),
    axis.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    x = "Environmental Change Magnitude",
    y = "Genomic Offset (ALL SNPs)",
    title = "Offset vs Environmental Change - Mean of All Models"
  )
print(p1)
```


























