---
title: "PRISM_RDA"
output: html_document
date: "2026-06-04"
---
# =============================================================================
# RDA-based Genomic Offset — Latent Space Approach
# =============================================================================
# Mirrors the PLS offset script. Key differences:
#   - Uses vegan::rda() LC scores instead of sPLS X variates
#   - Offset computed on significant RDA axes only (from permutation test)
#   - predict.rda(type = "lc") for future projection
#
# Assumes in environment:
#   - rda_mod            : fitted vegan rda object on historical data
#   - Y_2023             : allele freq matrix, pop × SNP
#   - X_hist             : scaled historical climate matrix (with scale attrs)
#   - hist_center/scale  : from attr(X_hist, "scaled:center/scale")
#   - pop_names          : character vector
#   - metadata           : Site_code, Elevation, Lat, Long
#   - climate_data_future: LOCA2 future TSV
#   - uncertainty_df     : from PLS script (vulnerability_class per pop)
# =============================================================================


```{r}
library(tidyverse)
library(vegan)
library(ggrepel)
library(patchwork)
library(ggthemes)
```

```{r}
# We would be using this new data and not the old one.
geno_23 = read.delim("C:/Users/jhana/OneDrive/Desktop/STTO_Simluation-main/STTO_Simluation/data/snp_allele_freq_mat.csv", header = TRUE, sep = ",")
pops =colnames(geno_23)[-1]
npop = length(pops)
```



#Read historical data
```{r}
his_data = read.delim("C:/Users/jhana/OneDrive/Desktop/STTO_Simluation-main/STTO_Simluation/data/PRISM_historical_data.tsv", header = TRUE)
```

#Read future data
```{r}
fut_data = read.delim("C:/Users/jhana/OneDrive/Desktop/STTO_Simluation-main/STTO_Simluation/data/LOCA2_climate_data.tsv", header = TRUE)
```


```{r}
ENV_VARS   <- c("ppt_mean", "ppt_cv", "tasmax_mean", "tasmax_cv", "tasmin_mean", 
"tasmin_cv")
MODELS     <- c("access-cm2", "cnrm-esm2-1", "ec-earth3-veg",
                "miroc6", "mpi-esm1-2-hr", "mri-esm2-0")
SCENARIOS  <- c("ssp245", "ssp370")
OUT_DIR    <- "results/rda_offset"

metadata = read_tsv("data/stto_pop_metadata.tsv")

dir.create(OUT_DIR, recursive = TRUE, showWarnings = FALSE)
```



# Step 1. Calculate the RDA
```{r}
allele_data = read_csv("data/snp_allele_freq_mat.csv")
SNP_data = allele_data$SNP_ID
allele_data = allele_data |>
select(-SNP_ID)
fams = colnames(allele_data)


X <- read_tsv("data/Flint_agg_data.tsv") |>
filter(population %in% fams) |>
pivot_longer(cols = ppt:tasmin, names_to = "cvar", values_to = "val") |>
group_by(population, cvar) |>
summarize(mean = mean(val),
cv = sd(val)/mean, .groups = "drop") |>
pivot_longer(cols = mean:cv) |>
unite(name, cvar:name) |>
pivot_wider(id_cols = population, names_from = name, values_from = value) |>
dplyr::select(population, all_of(ENV_VARS)) %>%
{ m <- as.matrix(.[, -1]); rownames(m) <- .$population; scale(m) }

hist_center <- attr(X, "scaled:center")
hist_scale  <- attr(X, "scaled:scale")

allele_freq_t = t(allele_data[,rownames(X)])

rda_obj = rda(allele_freq_t ~ ppt_mean + tasmax_mean + tasmin_mean + ppt_cv + tasmax_cv + tasmin_cv, 
    data = as.data.frame(X))

```

# =============================================================================
# SECTION 1 — Determine significant RDA axes
# =============================================================================
# Use your existing permutation test to decide how many axes to retain.
# If you've already run this, just set N_RDA_AXES manually.

#cat("Testing RDA axis significance via permutation...\n")
#axis_test <- anova(rda_obj, by = "axis", permutations = 100)
#print(axis_test)


```{r}
# Retain axes with p < 0.05
#sig_axes <- which(axis_test$`Pr(>F)` < 0.05)
N_RDA_AXES <- 2
cat("Significant RDA axes:", N_RDA_AXES, "\n")

# Variance explained by each axis
rda_varexp <- summary(eigenvals(rda_obj, constrained = TRUE))
cat("Variance explained by constrained axes:\n")
print(round(rda_varexp, 3))

# Historical LC scores for significant axes only
hist_lc <- scores(rda_obj, display = "lc", scaling = 1,
                  choices = 1:N_RDA_AXES)
# Ensure row order matches pop_names
hist_lc <- hist_lc[fams, , drop = FALSE]
```

# =============================================================================
# SECTION 2 — Future climate aggregation
# (reuse helpers from PLS script or redefine here)
# =============================================================================
```{r}
aggregate_future_climate <- function(future_df, model_id, scenario_id, env_vars,
                                      hist_center, hist_scale, pop_names) {
  df <- future_df |>
    filter(model == model_id,
           experiment == scenario_id) |>
    pivot_longer(cols = tasmax:ppt, names_to = "variable", values_to = "value") |>
group_by(population, variable) |>
summarize(mean = mean(value),
cv = sd(value)/mean, .groups = "drop") |>
pivot_longer(cols = mean:cv) |>
unite(name, variable:name) |>
pivot_wider(id_cols = population, names_from = name, values_from = value) |>
    filter(population %in% pop_names) |>
    arrange(match(population, pop_names)) |>
    column_to_rownames("population") |>
    as.matrix()

  # Scale using HISTORICAL center/scale — critical for valid projection
  scale(df, center = hist_center[env_vars], scale = hist_scale[env_vars])
}
```

```{r}
# Also build ensemble mean per scenario (average across models)
aggregate_ensemble_future <- function(future_df, scenario_id, env_vars,
                                       hist_center, hist_scale, pop_names) {
  future_df |>
    filter(experiment == scenario_id) |>
    group_by(population, year) |>
    summarise(across(tasmax:ppt, mean), .groups = "drop") |>
    pivot_longer(cols = tasmax:ppt, names_to = "variable", values_to = "value") |>
group_by(population, variable) |>
summarize(mean = mean(value),
cv = sd(value)/mean, .groups = "drop") |>
pivot_longer(cols = mean:cv) |>
unite(name, variable:name) |>
pivot_wider(id_cols = population, names_from = name, values_from = value)  |>
    filter(population %in% pop_names) |>
    arrange(match(population, pop_names)) |>
    column_to_rownames("population") |>
    as.matrix() |>
    scale(center = hist_center[env_vars], scale = hist_scale[env_vars])
}
```


```{r}
# =============================================================================
# SECTION 3 — Compute RDA offset for all model × scenario combinations
# =============================================================================

compute_rda_offset <- function(rda_model, X_future_scaled,
                                hist_lc, pop_names, n_axes) {

  # newdata must be a data frame with same predictor names as RDA formula
  newdata_df <- as.data.frame(X_future_scaled)

  # predict type = "lc": linear combination (fitted) site scores
  # these are purely constraint-driven, no residual WA component
  fut_lc <- predict(rda_model,
                    newdata = newdata_df,
                    type    = "lc",
                    scaling = 1)[ , 1:n_axes, drop = FALSE]

  shared_pops <- intersect(pop_names, rownames(fut_lc))
  hist_v <- hist_lc[shared_pops, 1:n_axes, drop = FALSE]
  fut_v  <- fut_lc[shared_pops,  1:n_axes, drop = FALSE]

  offset <- sqrt(rowSums((fut_v - hist_v)^2))

  list(
    offset      = offset,
    hist_lc     = hist_v,
    fut_lc      = fut_v,
    shared_pops = shared_pops
  )
}
```


```{r}
# ── Run all 12 + ensemble ──────────────────────────────────────────────────

climate_data_future = read_tsv("data/LOCA2_climate_data.tsv")
combo_grid <- expand_grid(model = MODELS, scenario = SCENARIOS)

cat("Computing RDA offsets...\n")

all_rda_offsets <- pmap_dfr(combo_grid, function(model, scenario) {
  cat(" →", model, "|", scenario, "\n")
  X_fut <- aggregate_future_climate(
    climate_data_future, model, scenario, ENV_VARS, hist_center, hist_scale, fams
  )
  res <- compute_rda_offset(rda_obj, X_fut, hist_lc, fams, N_RDA_AXES)
  data.frame(population = res$shared_pops, model, scenario, offset = res$offset)
})

ensemble_rda <- map_dfr(SCENARIOS, function(scenario) {
  X_ens <- aggregate_ensemble_future(
    climate_data_future, scenario, ENV_VARS, hist_center, hist_scale, fams
  )
  res <- compute_rda_offset(rda_obj, X_ens, hist_lc, fams, N_RDA_AXES)
  data.frame(population = res$shared_pops, model = "ensemble", scenario,
             offset = res$offset)
})

all_rda_offsets <- bind_rows(all_rda_offsets, ensemble_rda) |>
  left_join(metadata |> select(Site_code, Elevation, Lat, Long),
            by = c("population" = "Site_code"))

write_tsv(all_rda_offsets, file.path(OUT_DIR, "all_rda_offset_results.tsv"))
```
# =============================================================================
# SECTION 4 — Uncertainty summary
# =============================================================================

```{r}
rda_uncertainty_df <- all_rda_offsets |>
  filter(model != "ensemble") |>
  group_by(population, Elevation) |>
  summarise(
    offset_mean = mean(offset),
    offset_sd   = sd(offset),
    offset_cv   = sd(offset) / mean(offset),
    offset_min  = min(offset),
    offset_max  = max(offset),
    .groups     = "drop"
  )

med_offset_rda <- median(rda_uncertainty_df$offset_mean)
med_cv_rda     <- median(rda_uncertainty_df$offset_cv)

rda_uncertainty_df <- rda_uncertainty_df |>
  mutate(
    vulnerability_class = case_when(
      offset_mean >= med_offset_rda & offset_cv <  med_cv_rda ~ "Robustly vulnerable",
      offset_mean >= med_offset_rda & offset_cv >= med_cv_rda ~ "Vulnerable, uncertain",
      offset_mean <  med_offset_rda & offset_cv <  med_cv_rda ~ "Robustly buffered",
      offset_mean <  med_offset_rda & offset_cv >= med_cv_rda ~ "Genuinely uncertain"
    )
  )

vuln_colors <- c(
  "Robustly vulnerable"   = "#d73027",
  "Vulnerable, uncertain" = "#fc8d59",
  "Robustly buffered"     = "#1a9850",
  "Genuinely uncertain"   = "#878787"
)
```



# =============================================================================
# SECTION 5 — RDA biplot with future projections
# This is the key visualization unique to RDA: you can show the SNP loadings
# (species scores) alongside population trajectories in the same space
# =============================================================================

```{r}
build_rda_arrow_df <- function(rda_model, X_future_scaled,
                                pop_names, n_axes, model_id, scenario_id) {
  newdata_df <- as.data.frame(X_future_scaled)
  fut_lc <- predict(rda_model, newdata = newdata_df,
                    type = "lc", scaling = 1)[, 1:n_axes, drop = FALSE]
  hist_lc_plot <- scores(rda_model, display = "lc",
                         scaling = 1, choices = 1:n_axes)[pop_names, ]
  shared <- intersect(pop_names, rownames(fut_lc))

  data.frame(
    population = shared,
    x_hist     = hist_lc_plot[shared, 1],
    y_hist     = hist_lc_plot[shared, 2],
    x_fut      = fut_lc[shared, 1],
    y_fut      = fut_lc[shared, 2],
    model      = model_id,
    scenario   = scenario_id
  ) |>
    mutate(rda_dist = sqrt((x_fut - x_hist)^2 + (y_fut - y_hist)^2))
}

# Environmental variable arrows (biplot scores)
env_arrows <- scores(rda_obj, display = "bp", scaling = 1) |>
  as.data.frame() |>
  rownames_to_column("variable") |>
  rename(RDA1_arrow = RDA1, RDA2_arrow = RDA2)

# SNP loadings — top loaded SNPs on RDA1 and RDA2 for annotation
snp_scores <- scores(rda_obj, display = "species", scaling = 1) |>
  as.data.frame() |>
  rownames_to_column("snp") |>
  mutate(loading = sqrt(RDA1^2 + RDA2^2)) |>
  slice_max(loading, n = 20)   # top 20 SNPs by total loading

# Build arrow data for all combos
all_arrow_df <- pmap_dfr(combo_grid, function(model, scenario) {
  X_fut <- aggregate_future_climate(
    climate_data_future, model, scenario, ENV_VARS, hist_center, hist_scale, fams
  )
  build_rda_arrow_df(rda_obj, X_fut, fams, N_RDA_AXES, model, scenario)
}) |>
  left_join(rda_uncertainty_df |> select(population, vulnerability_class),
            by = "population")

# Axis labels
pct_rda <- round(summary(eigenvals(rda_obj, constrained = TRUE))[2, 1:2] * 100, 1)

# ── Plot: RDA biplot faceted by model × scenario ─────────────────────────────

arrow_scale <- 1.5   # scale env arrows for visibility

p_rda_biplot <- ggplot() +
  # SNP scores — light background cloud
  geom_point(data = scores(rda_obj, display = "species", scaling = 1) |>
               as.data.frame(),
             aes(x = RDA1, y = RDA2),
             color = "grey85", size = 0.4, alpha = 0.5) +
  # Population displacement arrows
  geom_segment(
    data = all_arrow_df,
    aes(x = x_hist, y = y_hist, xend = x_fut, yend = y_fut,
        color = rda_dist),
    arrow     = arrow(length = unit(0.15, "cm"), type = "closed"),
    linewidth = 0.6, alpha = 0.85
  ) +
  # Historical positions
  geom_point(data = all_arrow_df,
             aes(x = x_hist, y = y_hist),
             shape = 21, fill = "white", color = "grey30", size = 2) +
  geom_text_repel(
    data = all_arrow_df |> distinct(population, x_hist, y_hist),
    aes(x = x_hist, y = y_hist, label = population),
    size = 2, color = "grey20", max.overlaps = 12
  ) +
  # Environmental variable arrows
  geom_segment(
    data = env_arrows,
    aes(x = 0, y = 0,
        xend = RDA1_arrow * arrow_scale,
        yend = RDA2_arrow * arrow_scale),
    arrow     = arrow(length = unit(0.2, "cm"), type = "closed"),
    color     = "steelblue", linewidth = 0.8
  ) +
  geom_text(
    data = env_arrows,
    aes(x = RDA1_arrow * arrow_scale * 1.15,
        y = RDA2_arrow * arrow_scale * 1.15,
        label = variable),
    color = "steelblue", size = 3, fontface = "bold"
  ) +
  scale_color_viridis_c(option = "magma", direction = -1,
                         name = "RDA\ndistance") +
  facet_grid(scenario ~ model) +
  theme_bw(base_size = 8) +
  theme(
    strip.background = element_rect(fill = "grey92"),
    strip.text       = element_text(face = "bold", size = 7),
    legend.position  = "right"
  ) +
  labs(
    x     = paste0("RDA1 (", pct_rda[1], "% constrained var)"),
    y     = paste0("RDA2 (", pct_rda[2], "% constrained var)"),
    title = "Population displacement in RDA space — historical → future",
    subtitle = "Grey cloud = SNP loadings; blue arrows = env variables; colored arrows = population trajectories"
  )

ggsave(file.path(OUT_DIR, "rda_biplot_all_scenarios.png"),
       p_rda_biplot, width = 22, height = 8, limitsize = FALSE)


cat("\n=== RDA offset complete ===\n")
cat("Outputs in:", OUT_DIR, "\n")
```


