---
title: "Example code"
author: "Rishav Ray"
output: pdf_document
date: "2025-10-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Example code for LFMM analysis

Let's load up the libraries and codes

```{r}
library(tidyverse)
library(ggthemes)
library(ggrepel)
library(patchwork)
library(lfmm)
source("manhattan_plot.R")
```

Load the frequency file, and other data such as climate data.
There are two files in the data folder, a historic and future projection.
Load the two files and take a look at it and see how they are organized.

```{r}
geno_23 = read.delim("data/snp_allele_freq_mat.csv", header = TRUE, sep =",")

# I'm going to simulate some climate data here. But ideally they should be real data
pops =colnames(geno_23)[-1]
npop = length(pops)
climate_data_sim = tibble(PopID = pops,
max_temp = rnorm(npop),
min_temp = rnorm(npop),
cwd = rnorm(npop),
ppt = rnorm(npop))

loo_cv = readRDS("data/loo_cv.RDS")
```


Now set up the data for LFMM

```{r}
Y_2023 = as.matrix(geno_23[,-1])
Y_2023 = t(Y_2023)

# I would not scale the climate data here, since it is already with mean = 0 and sd = 1
# But it should be scaled when using real data
# X_2023 = scale(X_2023)

X_2023 = as.matrix(climate_data_sim[,-1])
rownames(X_2023) = pops

# We align the pop codes in the both X and Y matrix
common = intersect(rownames(X_2023), rownames(Y_2023))
Y_2023 = Y_2023[common,]
X_2023 = X_2023[common, , drop = FALSE]

```

And we run the LFMM and then calculate the p-values

```{r}
mod.lfmm = lfmm_lasso(Y = Y_2023, 
    X = X_2023, 
    K = 5)

pv_2023 = lfmm_test(Y = Y_2023, 
    X = X_2023, 
    lfmm = mod.lfmm, 
    calibrate = "gif")
```

Make some qqplot

```{r}
#I'm doing the first p-value col, that is the max_temp
# This should be done for all the climate variables. Ideally make a function for this

pvalues_2023.raw = pv_2023$pvalue[,1]

n = length(sort(pvalues_2023.raw))
df = tibble(observed = sort(-log10(pvalues_2023.raw)),
expected = sort(-log10(ppoints(n))),
clower   = -log10(qbeta(p = (1 - 0.95) / 2, shape1 = 1:n, shape2 = n:1)),
cupper   = -log10(qbeta(p = (1 + 0.95) / 2, shape1 = 1:n, shape2 = n:1))
)

# Convert p-values to 1-df chi-squared statistics
chi2 <- qchisq(pvalues_2023.raw, df=1, lower.tail=FALSE)
# Calculate lambda
lambda <- median(chi2) / qchisq(0.5, df=1)
print(lambda)


df |>
ggplot() +
geom_ribbon(aes(x = expected, ymin = clower, ymax = cupper), alpha = 0.1) +
geom_point(aes(expected, observed), shape = 1, size = 3) +
geom_abline(intercept = 0, slope = 1, alpha = 0.5) +
theme_bw() +
labs(x = expression(paste("Expected -log"[10], plain(P))),
y = expression(paste("Observed -log"[10], plain(P))))
```

Finally make some manhattan plot

```{r}
plot_data_2023 = tibble(loci = geno_23[,1], p_val = pv_2023$pvalue[,1], calibrated.pvalue = pv_2023$calibrated.pvalue[,1]) |>
separate_wider_delim(loci, delim = "_", names = c("chr", "bp")) |>
filter(str_detect(chr, "Chr")) |>
mutate(p.bonf = p.adjust(p_val, method = "bonferroni"),
bp = as.numeric(bp))


manhattan_WL2 = plot_manhattan(plot_data_2023,
                        chr_col = "chr",
                        pos_col = "bp",
                        value_col = "calibrated.pvalue",
                        transform = "neglog10",
                        title = "WL2 2023",
                        highlight_top_n = 20)

print(manhattan_WL2)

# Optionally save plot
#ggsave("WL2_2023.png", manhattan_WL2, width = 12, height = 6, dpi = 300)
```

To get the genes in top hits, we select top 20 loci. Ideally we aim for 0.01%, so play around with the cutoff.

```{r}
top20 = plot_data |>
slice_min(calibrated.pvalue, n = 20) |>
select(seqid = chr, bp, p_val, calibrated.pvalue, p.bonf)
write_tsv(top20_UCD, "Data/tables/top20_loc.tsv")
```

Now join that with the genes data and extract genes in a certain window.

```{r}
genes = read_tsv("data/gene_description.tsv")
window_size = 10000
```

Now run the code to find nearby genes for UCD 2022

```{r}
genes_near_snps = top20 |>
  left_join(genes, by = "seqid", relationship = "many-to-many") |>
  mutate(
    # Calculate distances
    dist_to_start = bp - start,
    dist_to_end = bp - end,
    
    # Check if SNP is within gene body
    in_gene = bp >= start & bp <= end,
    
    # Check if SNP is within window (upstream or downstream)
    in_window = (bp >= (start - window_size) & bp <= (end + window_size))
  ) |>
  filter(in_window) |>
  mutate(
    # Categorize position
    position = case_when(
      in_gene ~ "in_gene",
      bp < start ~ "upstream",
      bp > end ~ "downstream"
    ),
    distance = case_when(
      in_gene ~ 0,
      bp < start ~ start - bp,
      bp > end ~ bp - end
    )
  ) |>
  arrange(seqid, bp, distance)

write_tsv(genes_near_snps_UCD, "Data_tables/genes_10kb.tsv")
```

## Calculating genomic offset

We first run LFMM on the historic data. And subsequently project it back to future climate data and calculate the distance.

```{r}
geno_23 = read.delim("data/merged_maf_common_WL2_2023.tsv", header = TRUE)

# I'm going to simulate some climate data here. But ideally they should be real data
pops =colnames(geno_23)[-1]
npop = length(pops)

climate_data_sim_historic = tibble(PopID = pops,
max_temp = rnorm(npop),
min_temp = rnorm(npop),
ppt = rnorm(npop))

climate_data_sim_future = tibble(PopID = pops,
max_temp = rnorm(npop),
min_temp = rnorm(npop),
ppt = rnorm(npop))

```

Run LFMM

```{r}
Y_2023 = as.matrix(geno_23[,-1])
Y_2023 = t(Y_2023)

# I would not scale the climate data here, since it is already with mean = 0 and sd = 1
# But it should be scaled when using real data
# X_2023 = scale(X_2023)

X_2023 = as.matrix(climate_data_sim_historic[,-1])
rownames(X_2023) = pops

# We align the pop codes in the both X and Y matrix
common = intersect(rownames(X_2023), rownames(Y_2023))
Y_2023 = Y_2023[common,]
X_2023 = X_2023[common, , drop = FALSE]

```


```{r}
mod.lfmm = lfmm_ridge(Y = Y_2023, 
    X = X_2023, 
    K = 2)

pv_2023 = lfmm_test(Y = Y_2023, 
    X = X_2023, 
    lfmm = mod.lfmm, 
    calibrate = "gif")
```

Make a function to calculate genomic offset 

```{r}
calculate_genomic_offset = function(allele_freq, 
                                     lfmm_effects, 
                                     env_current, 
                                     env_future,
                                     snp_filter = NULL,
                                     method = "euclidean") {
  
  # Dimensions check
  n_pop = nrow(allele_freq)
  n_snp = ncol(allele_freq)
  n_env = ncol(env_current)
  
  cat("\n=== Genomic Offset Calculation ===\n")
  cat("Populations:", n_pop, "\n")
  cat("SNPs (total):", n_snp, "\n")
  cat("Environmental variables:", n_env, "\n")
  
  # Apply SNP filter if provided
  if (!is.null(snp_filter)) {
    allele_freq = allele_freq[, snp_filter, drop = FALSE]
    lfmm_effects = lfmm_effects[snp_filter, , drop = FALSE]
    cat("SNPs (after filtering):", sum(snp_filter), "\n")
  }
  
  # Ensure environmental variables are standardized
  # Use same scaling for current and future based on current
  env_means = colMeans(env_current)
  env_sds = apply(env_current, 2, sd)
  
  #env_current_scaled = scale(env_current, center = env_means, scale = env_sds)
  #env_future_scaled = scale(env_future, center = env_means, scale = env_sds)
  env_current_scaled = env_current
  env_future_scaled = env_future
  
  cat("\nEnvironmental variables standardized using current climate statistics\n")
  
  # ========================================================================
  # STEP 1: Calculate genetic values under current climate
  # ========================================================================
  # For each population, calculate adaptive genetic value
  # genetic_value = sum over SNPs and env vars of:
  #   (allele_freq × effect_size × env_value)
  
  genetic_values_current = matrix(0, nrow = n_pop, ncol = n_env)
  
  for (i in 1:n_pop) {
    for (j in 1:n_env) {
      # Allele frequency × effect size for this env variable
      weighted_freqs = allele_freq[i, ] * lfmm_effects[, j]
      # Weighted by environmental value
      genetic_values_current[i, j] = sum(weighted_freqs) * env_current_scaled[i, j]
    }
  }
  
  # Total genetic value across all environmental axes
  genetic_value_current_total = rowSums(genetic_values_current)
  
  cat("\nCurrent genetic values calculated\n")
  
  # ========================================================================
  # STEP 2: Calculate genetic values under future climate
  # ========================================================================
  
  genetic_values_future = matrix(0, nrow = n_pop, ncol = n_env)
  
  for (i in 1:n_pop) {
    for (j in 1:n_env) {
      # Same allele frequencies (no evolution assumed)
      # But different environmental values
      weighted_freqs = allele_freq[i, ] * lfmm_effects[, j]
      genetic_values_future[i, j] = sum(weighted_freqs) * env_future_scaled[i, j]
    }
  }
  
  genetic_value_future_total = rowSums(genetic_values_future)
  
  cat("Future genetic values calculated\n")
  
  # ========================================================================
  # STEP 3: Calculate offset
  # ========================================================================
  
  if (method == "euclidean") {
    # Euclidean distance across all environmental dimensions
    offset = sqrt(rowSums((genetic_values_future - genetic_values_current)^2))
    cat("\nOffset method: Euclidean distance\n")
    
  } else if (method == "absolute") {
    # Simple absolute difference
    offset = abs(genetic_value_future_total - genetic_value_current_total)
    cat("\nOffset method: Absolute difference\n")
    
  } else if (method == "relative") {
    # Relative change
    offset = abs((genetic_value_future_total - genetic_value_current_total) / 
                    (abs(genetic_value_current_total) + 1e-10))
    cat("\nOffset method: Relative change\n")
  }
  
  # ========================================================================
  # STEP 4: Per-variable contributions
  # ========================================================================
  
  # Calculate offset contribution from each environmental variable
  offset_by_var = sqrt((genetic_values_future - genetic_values_current)^2)
  colnames(offset_by_var) = colnames(env_current)
  
  cat("\nOffset calculation complete!\n")
  cat("Mean offset:", round(mean(offset), 4), "\n")
  cat("Range:", round(min(offset), 4), "-", round(max(offset), 4), "\n")
  
  # Return results
  return(list(
    offset = offset,
    offset_by_variable = offset_by_var,
    genetic_values_current = genetic_value_current_total,
    genetic_values_future = genetic_value_future_total,
    genetic_values_current_by_var = genetic_values_current,
    genetic_values_future_by_var = genetic_values_future,
    env_change = env_future_scaled - env_current_scaled,
    n_snps_used = ncol(allele_freq)
  ))
}
```

Calculate the offset now


```{r}
allele_freq = Y_2023
env_current = as.matrix(climate_data_sim_historic[,-1])
rownames(env_current) = pops
env_future = as.matrix(climate_data_sim_future[,-1])
rownames(env_future) = pops
lfmm_effects = mod.lfmm$B
# lfmm_pvalues = pv_2023$pvalue
pop_names = rownames(allele_freq)


offset_all = calculate_genomic_offset(
  allele_freq = allele_freq,
  lfmm_effects = lfmm_effects,
  env_current = env_current,
  env_future = env_future,
  snp_filter = NULL,  # Use all SNPs
  method = "euclidean"
)
```

Plot results (basic)

```{r}
results_df <- data.frame(
  population = pop_names,
  # Offsets
  offset_all_snps = offset_all$offset,
  # Genetic values
  genetic_value_current_all = offset_all$genetic_values_current,
  genetic_value_future_all = offset_all$genetic_values_future,
  # Add environmental change magnitude
  env_change_magnitude = sqrt(rowSums((env_future - env_current)^2))
)
```

```{r}
p1 <- ggplot(results_df, aes(x = env_change_magnitude, y = offset_all_snps, label = population)) +
  geom_point(size = 3, alpha = 0.7, color = "darkred") +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  geom_text_repel()+
  theme_minimal() +
  theme(
    axis.text = element_text(size = 11),
    axis.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    x = "Environmental Change Magnitude",
    y = "Genomic Offset (ALL SNPs)",
    title = "Offset vs Environmental Change"
  )
print(p1)
```

## Make a map of the populations

```{r}
library(maps) # Required to download and plot the map
library(tidyverse)
library(ggthemes)
library(ggdoctheme)
library(ggrepel)

ca_map = map_data("state") |>
  filter(region == "california")

# Read the coordinates and the metadata of the populations
stto_data = read_tsv("data/stto_metadata.tsv")

# Define the population Site_codes to retain
pops = c("BH", "CC", "CP2", "CP3", "DPR", "FR", "IH", "LV1", "LV3", 
"LVTR1", "SC", "SQ1", "SQ2", "SQ3", "TM2", "WL1", "WL2", "WR", 
"YO4", "YO7", "YO8", "YO11")

# create the plotdata dataframe

plotdata = data.frame(Site_code = pops) |>
left_join(stto_data) |>
dplyr::select(Site_code, Lat,	Long,	Elevation, Label)


ggplot() +
  # Plot California base map
  geom_polygon(data = ca_map, 
               aes(x = long, y = lat, group = group),
               fill = "white", color = "black") +
  # Add points for locations
  geom_point(data = plotdata,
             aes(x = Long, y = Lat),
             alpha = 0.7, size = 3, position = position_dodge(width = 1)) +
  # Add labels for locations
  geom_text_repel(data = plotdata,
            aes(x = Long, y = Lat, label = Site_code),
            size = 6) +
  # Customize the appearance
  # Set the coordinate system and zoom
  coord_fixed(1.3) +
  # Customize the theme
  theme_void()

```

#New Allele Frequency matrix with alpha = 0.3
```{r}
alpha <- 0.7

new_LV1_03 = ((1 - alpha) * geno_23$LV1) + (alpha * geno_23$SQ1)
new_allele_mat_03 <- geno_23
new_allele_mat_03$LV1 = new_LV1_03
```

```{r}
plot(new_allele_mat_03$LV1, geno_23$LV1, xlab = "new LV1", ylab = "old LV1", main = "Admixture Plot alpha = 0.3")
```

#offset calculations
```{r}
#admixed offset
# Extract B here
B = pv_lasso_noLV1$mod$B
adx_allele_freq_03 = as.matrix(new_allele_mat_03[, -1]) #remove snpid column
adx_allele_freq_03 = t(adx_allele_freq_03)

# Align to common populations with env data
common_adx_03 = intersect(
  intersect(rownames(X_2023_gcm_mean), rownames(X_gcm_fut)),
  rownames(adx_allele_freq_03)
)

adx_allele_freq_03 = adx_allele_freq_03[common_adx_03, , drop = FALSE]
env_current_03 = as.matrix(X_2023_gcm_mean[common_adx_03, , drop = FALSE])
#rownames(env_current) = rownames(X_2023_gcm_mean)
env_future_03 = as.matrix(X_gcm_fut[common_adx_03, , drop = FALSE])
#rownames(env_future) = rownames(X_gcm_fut)
lfmm_effects = B
# lfmm_pvalues = pv_2023$pvalue




adx_offset_all_gcm_mean_03 = calculate_genomic_offset(
  allele_freq = adx_allele_freq_03,
  lfmm_effects = B,
  env_current = env_current_03,
  env_future = env_future_03,
  snp_filter = NULL,  # Use all SNPs
  method = "euclidean"
)
```

```{r}
results_df_adx_03 <- data.frame(
  population = pop_names,
  # Offsets
  offset_all_snps = adx_offset_all_gcm_mean_03$offset,
  # Genetic values
  genetic_value_current_all = adx_offset_all_gcm_mean_03$genetic_values_current,
  genetic_value_future_all = adx_offset_all_gcm_mean_03$genetic_values_future,
  # Add environmental change magnitude
  env_change_magnitude = sqrt(rowSums((env_future - env_current)^2))
)
```

```{r}
p77 <- ggplot(results_df_adx_03, aes(x = env_change_magnitude, y = offset_all_snps, label = population)) +
  geom_point(size = 3, alpha = 0.7, color = "darkred") +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  geom_text_repel()+
  theme_minimal() +
  theme(
    axis.text = element_text(size = 11),
    axis.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    x = "Environmental Change Magnitude",
    y = "Genomic Offset (ALL SNPs)",
    title = "Offset vs Environmental Change - Mean of All Models Admixed Mixture alpha = 0.3"
  )
print(p77)
```

Some diagnostics

```{r}
# Just simple check on pop frequency times B matrix per environmental variable
B = loo_cv$full_model$B

bfreq_LV1 <- Y_2023["LV1", ] %*% B   # 1 × n_env
bfreq_SQ3 <- Y_2023["SQ3", ] %*% B   # 1 × n_env

# They are opposite
rbind(LV1 = bfreq_LV1, SQ3 = bfreq_SQ3)

# Direction of environmental change for LV1 and SQ3
delta_X_LV1 <- env_future["LV1", ] - env_current["LV1", ]
print(delta_X_LV1)

delta_X_SQ3 <- env_future["SQ3", ] - env_current["SQ3", ]
print(delta_X_SQ3)

# The genetic value shift per variable is bfreq × delta_X
# If bfreq_LV1 and delta_X are the same sign = positive offset contribution
# Testing whether introducing SQ3 alleles changes bfreq — does it move it toward or away
# from cancelling delta_X?

gval_shift_LV1 <- as.numeric(bfreq_LV1) * as.numeric(delta_X_LV1) 
gval_shift_SQ3 <- as.numeric(bfreq_SQ3) * as.numeric(delta_X_LV1)

# The sign of gval_shift tells us: 
# positive = this variable increases offset
# negative = this variable decreases offset
# If SQ3's contribution has the same or larger sign than LV1's, then admixture will not help

rbind(
  LV1_gval_shift = gval_shift_LV1,
  SQ3_gval_shift = gval_shift_SQ3
) %>% 
  as.data.frame() %>%
  setNames(colnames(env_current)) %>%
  print()
```

```{r}
simulate_rescue_generations <- function(Y_full,
                                         recipient,
                                         donor,
                                         B_loo,
                                         X_current,
                                         X_future,
                                         alpha        = 0.20,
                                         n_gen        = 10,
                                         n_ind        = 200,
                                         selection    = TRUE) {
  
  n_snp <- ncol(Y_full)
  
  # ── Step 1: Sample individual genotypes from population frequencies ──────────
  # Under HWE, genotype counts are binomial draws from allele frequency
  # Returns n_ind × n_snp matrix of {0, 1, 2} genotypes
  
  sample_genotypes <- function(freq_vec, n) {
    matrix(rbinom(n * length(freq_vec), size = 2, prob = freq_vec),
           nrow = n, ncol = length(freq_vec), byrow = TRUE)
  }
  
  # ── Step 2: Compute individual genetic values and offset ─────────────────────
  # Individual genetic value = genotype_vec %*% B_loo (summed across env vars)
  # Offset = Euclidean distance in genetic value space current vs future
  
  compute_individual_offset <- function(geno_mat) {
    G_current <- (geno_mat %*% B_loo) * 
                  matrix(X_current[recipient, ], nrow = nrow(geno_mat), 
                         ncol = ncol(B_loo), byrow = TRUE)
    G_future  <- (geno_mat %*% B_loo) * 
                  matrix(X_future[recipient, ],  nrow = nrow(geno_mat), 
                         ncol = ncol(B_loo), byrow = TRUE)
    sqrt(rowSums((G_future - G_current)^2))   # n_ind offset values
  }
  
  # ── Step 3: Reproduce next generation with recombination ─────────────────────
  # Each offspring draws alleles from two randomly paired parents
  # Recombination: at each SNP, allele drawn from parent 1 or parent 2
  # with equal probability — approximates free recombination
  
  reproduce <- function(geno_mat, n_offspring) {
    n    <- nrow(geno_mat)
    # Random parent pairs
    p1   <- geno_mat[sample(n, n_offspring, replace = TRUE), ]
    p2   <- geno_mat[sample(n, n_offspring, replace = TRUE), ]
    # At each SNP, each allele drawn from p1 or p2 with prob 0.5
    mask <- matrix(rbinom(n_offspring * n_snp, 1, 0.5),
                   nrow = n_offspring, ncol = n_snp)
    p1 * mask + p2 * (1 - mask)
  }
  
  # ── Step 4: Soft selection ───────────────────────────────────────────────────
  # Individuals with lower offset under future climate have higher fitness
  # Fitness = 1 / (1 + offset) — simple inverse, no extra parameters
  # Sample next generation proportional to fitness
  
  select_individuals <- function(geno_mat) {
    offsets <- compute_individual_offset(geno_mat)
    fitness <- 1 / (1 + offsets)
    geno_mat[sample(nrow(geno_mat), nrow(geno_mat), replace = TRUE,
                    prob = fitness / sum(fitness)), ]
  }
  
  # ── Step 5: Population offset from allele frequencies ───────────────────────
  pop_offset_from_geno <- function(geno_mat) {
    freq <- colMeans(geno_mat) / 2   # convert {0,1,2} back to [0,1] frequency
    calculate_genomic_offset(
      allele_freq  = matrix(freq, nrow = 1),
      lfmm_effects = B_loo,
      env_current  = X_current[recipient, , drop = FALSE],
      env_future   = X_future[recipient, , drop = FALSE]
    )$offset
  }
  
  # ── Main simulation ──────────────────────────────────────────────────────────
  
  # Initialize populations
  geno_recipient <- sample_genotypes(Y_full[recipient, ], n_ind)
  geno_donor     <- sample_genotypes(Y_full[donor, ],     n_ind)
  
  # F0: admix at proportion alpha by pooling individuals
  n_donor_ind     <- round(alpha * n_ind)
  n_recipient_ind <- n_ind - n_donor_ind
  
  geno_current <- rbind(
    geno_recipient[1:n_recipient_ind, ],
    geno_donor[1:n_donor_ind, ]
  )
  
  # Record offset at each generation
  results <- data.frame(
    generation = 0:n_gen,
    offset     = NA_real_
  )
  results$offset[1] <- pop_offset_from_geno(geno_current)
  
  for (g in seq_len(n_gen)) {
    if (selection) geno_current <- select_individuals(geno_current)
    geno_current        <- reproduce(geno_current, n_ind)
    results$offset[g+1] <- pop_offset_from_geno(geno_current)
  }
  
  return(results)
}
```

```{r}
# Single donor-recipient pair
res <- simulate_rescue_generations(
  Y_full    = Y_2023,
  recipient = "LV1",
  donor     = "SQ1",
  B_loo     = B,
  X_current = env_current,
  X_future  = env_future,
  alpha     = 0.20,
  n_gen     = 10,
  n_ind     = 200,
  selection = TRUE
)

# Compare multiple donors
donors <- c("SQ1", "SQ2", "SQ3", "BH")

all_res <- lapply(donors, function(d) {
  simulate_rescue_generations(
    Y_full = Y_2023, recipient = "LV1", donor = d,
    B_loo = B, X_current = env_current, 
    X_future = env_future, alpha = 0.20, n_gen = 10
  ) %>% mutate(donor = d)
}) %>% bind_rows()

# Baseline — no admixture
baseline <- simulate_rescue_generations(
  Y_full = Y_2023, recipient = "LV1", donor = "LV1",
  B_loo = B, X_current = env_current,
  X_future = env_future, alpha = 0, n_gen = 10
) %>% mutate(donor = "No admixture")

# Plot
bind_rows(all_res, baseline) %>%
  ggplot(aes(x = generation, y = offset, 
             color = donor, group = donor)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  theme_doc() +
  labs(x = "Generation", y = "Genomic Offset",
       color = "Donor",
       title = paste("Simulated rescue trajectory — recipient: LV1"),
       subtitle = "Selection on offset each generation; recombination via free assortment")
```

RDA analysis sample code

```{r}
# Initialize the vegan package and source the plotting script
library(vegan)
library(tidyverse)
source("analysis_functions.R")
```

```{r}
fams = scan("data/fam.txt", what = character())

climate_data = read_tsv("data/Historical_climate_caladapt_1950_2004.tsv") |>
filter(id %in% fams) |>
group_by(id, cvar, dt) |>
summarize(mean_val = mean(val)) |>
ungroup()

aggregate_data = climate_data |>
group_by(id, cvar, dt) |>
filter(dt >= '1990-12-31') |>
summarize(mean = mean(mean_val)) |>
cv = sd(mean_val)/mean) |>
ungroup() |>
pivot_longer(cols = mean:cv) |>
unite(name, cvar:name) |>
pivot_wider(id_cols = id, names_from = name, values_from = value)

allele_data = read_csv("data/snp_allele_freq_mat.csv")
SNP_data = allele_data$SNP_ID
allele_data = allele_data |>
select(-SNP_ID)
allele_freq_t = t(allele_data[,aggregate_data$id])

```


```{r}
env_vars = aggregate_data[, c("pr_mean", "pr_cv", "tasmax_mean", "tasmax_cv", "tasmin_mean", 
"tasmin_cv")]

env_vars_scaled = scale(env_vars)

# Pure environmental model (IBE)
env_rda = rda(allele_freq_t ~ ., data = as.data.frame(env_vars_scaled))

env_rda_scores = extract_rda_scores(env_rda)
env_biplot = plot_rda_biplot(env_rda_scores, title = "Mean of all model", vector_scale = 2)

print(env_biplot)

```

```{r}
loo_cv   <- readRDS("data/loo_cv.RDS")

final_df <- loo_cv$comparison %>%
  left_join(metadata %>% select(Site_code, Elevation, Lat, Long),
            by = c("population" = "Site_code")) %>%
  mutate(
    leverage_class = case_when(
      offset_loo > offset_full * 1.5 ~ "Underestimated (full model)",
      offset_loo < offset_full * 0.5 ~ "Overestimated (full model)",
      TRUE                           ~ "Stable"
    ),
    cluster   = cluster_assignments[population],
    is_lv     = population %in% lv_pops,
    is_lever  = population %in% leverage_pops
  )

```