---
title: "Example code"
author: "Rishav Ray"
output: pdf_document
date: "2025-10-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Example code for LFMM analysis

Let's load up the libraries and codes

```{r}
library(tidyverse)
library(ggthemes)
library(ggrepel)
library(patchwork)
library(lfmm)
source("manhattan_plot.R")
```

Load the frequency file, and other data such as climate data.
There are two files in the data folder, a historic and future projection.
Load the two files and take a look at it and see how they are organized.

```{r}
geno_23 = read.delim("data/merged_maf_common_WL2_2023.tsv", header = TRUE)

# I'm going to simulate some climate data here. But ideally they should be real data
pops =colnames(geno_23)[-1]
npop = length(pops)
climate_data_sim = tibble(PopID = pops,
max_temp = rnorm(npop),
min_temp = rnorm(npop),
cwd = rnorm(npop),
ppt = rnorm(npop))
```

Now set up the data for LFMM

```{r}
Y_2023 = as.matrix(geno_23[,-1])
Y_2023 = t(Y_2023)

# I would not scale the climate data here, since it is already with mean = 0 and sd = 1
# But it should be scaled when using real data
# X_2023 = scale(X_2023)

X_2023 = as.matrix(climate_data_sim[,-1])
rownames(X_2023) = pops

# We align the pop codes in the both X and Y matrix
common = intersect(rownames(X_2023), rownames(Y_2023))
Y_2023 = Y_2023[common,]
X_2023 = X_2023[common, , drop = FALSE]

```

And we run the LFMM and then calculate the p-values

```{r}
mod.lfmm = lfmm_ridge(Y = Y_2023, 
    X = X_2023, 
    K = 2)

pv_2023 = lfmm_test(Y = Y_2023, 
    X = X_2023, 
    lfmm = mod.lfmm, 
    calibrate = "gif")
```

Make some qqplot

```{r}
#I'm doing the first p-value col, that is the max_temp
# This should be done for all the climate variables. Ideally make a function for this

pvalues_2023.raw = pv_2023$pvalue[,1]

n = length(sort(pvalues_2023.raw))
df = tibble(observed = sort(-log10(pvalues_2023.raw)),
expected = sort(-log10(ppoints(n))),
clower   = -log10(qbeta(p = (1 - 0.95) / 2, shape1 = 1:n, shape2 = n:1)),
cupper   = -log10(qbeta(p = (1 + 0.95) / 2, shape1 = 1:n, shape2 = n:1))
)

# Convert p-values to 1-df chi-squared statistics
chi2 <- qchisq(pvalues_2023.raw, df=1, lower.tail=FALSE)
# Calculate lambda
lambda <- median(chi2) / qchisq(0.5, df=1)
print(lambda)


df |>
ggplot() +
geom_ribbon(aes(x = expected, ymin = clower, ymax = cupper), alpha = 0.1) +
geom_point(aes(expected, observed), shape = 1, size = 3) +
geom_abline(intercept = 0, slope = 1, alpha = 0.5) +
theme_bw() +
labs(x = expression(paste("Expected -log"[10], plain(P))),
y = expression(paste("Observed -log"[10], plain(P))))
```

Finally make some manhattan plot

```{r}
plot_data_2023 = tibble(loci = geno_23[,1], p_val = pv_2023$pvalue[,1], calibrated.pvalue = pv_2023$calibrated.pvalue[,1]) |>
separate_wider_delim(loci, delim = ":", names = c("chr", "bp")) |>
filter(str_detect(chr, "Chr")) |>
mutate(p.bonf = p.adjust(p_val, method = "bonferroni"),
bp = as.numeric(bp))


manhattan_WL2 = plot_manhattan(plot_data_2023,
                        chr_col = "chr",
                        pos_col = "bp",
                        value_col = "calibrated.pvalue",
                        transform = "neglog10",
                        title = "WL2 2023",
                        highlight_top_n = 20)

print(manhattan_WL2)

# Optionally save plot
#ggsave("WL2_2023.png", manhattan_WL2, width = 12, height = 6, dpi = 300)
```

To get the genes in top hits, we select top 20 loci. Ideally we aim for 0.01%, so play around with the cutoff.

```{r}
top20 = plot_data |>
slice_min(calibrated.pvalue, n = 20) |>
select(seqid = chr, bp, p_val, calibrated.pvalue, p.bonf)
write_tsv(top20_UCD, "Data/tables/top20_loc.tsv")
```

Now join that with the genes data and extract genes in a certain window.

```{r}
genes = read_tsv("data/gene_description.tsv")
window_size = 10000
```

Now run the code to find nearby genes for UCD 2022

```{r}
genes_near_snps = top20 |>
  left_join(genes, by = "seqid", relationship = "many-to-many") |>
  mutate(
    # Calculate distances
    dist_to_start = bp - start,
    dist_to_end = bp - end,
    
    # Check if SNP is within gene body
    in_gene = bp >= start & bp <= end,
    
    # Check if SNP is within window (upstream or downstream)
    in_window = (bp >= (start - window_size) & bp <= (end + window_size))
  ) |>
  filter(in_window) |>
  mutate(
    # Categorize position
    position = case_when(
      in_gene ~ "in_gene",
      bp < start ~ "upstream",
      bp > end ~ "downstream"
    ),
    distance = case_when(
      in_gene ~ 0,
      bp < start ~ start - bp,
      bp > end ~ bp - end
    )
  ) |>
  arrange(seqid, bp, distance)

write_tsv(genes_near_snps_UCD, "Data_tables/genes_10kb.tsv")
```

## Calculating genomic offset

We first run LFMM on the historic data. And subsequently project it back to future climate data and calculate the distance.

```{r}
geno_23 = read.delim("data/merged_maf_common_WL2_2023.tsv", header = TRUE)

# I'm going to simulate some climate data here. But ideally they should be real data
pops =colnames(geno_23)[-1]
npop = length(pops)

climate_data_sim_historic = tibble(PopID = pops,
max_temp = rnorm(npop),
min_temp = rnorm(npop),
ppt = rnorm(npop))

climate_data_sim_future = tibble(PopID = pops,
max_temp = rnorm(npop),
min_temp = rnorm(npop),
ppt = rnorm(npop))

```

Run LFMM

```{r}
Y_2023 = as.matrix(geno_23[,-1])
Y_2023 = t(Y_2023)

# I would not scale the climate data here, since it is already with mean = 0 and sd = 1
# But it should be scaled when using real data
# X_2023 = scale(X_2023)

X_2023 = as.matrix(climate_data_sim_historic[,-1])
rownames(X_2023) = pops

# We align the pop codes in the both X and Y matrix
common = intersect(rownames(X_2023), rownames(Y_2023))
Y_2023 = Y_2023[common,]
X_2023 = X_2023[common, , drop = FALSE]

```


```{r}
mod.lfmm = lfmm_ridge(Y = Y_2023, 
    X = X_2023, 
    K = 2)

pv_2023 = lfmm_test(Y = Y_2023, 
    X = X_2023, 
    lfmm = mod.lfmm, 
    calibrate = "gif")
```

Make a function to calculate genomic offset 

```{r}
calculate_genomic_offset = function(allele_freq, 
                                     lfmm_effects, 
                                     env_current, 
                                     env_future,
                                     snp_filter = NULL,
                                     method = "euclidean") {
  
  # Dimensions check
  n_pop = nrow(allele_freq)
  n_snp = ncol(allele_freq)
  n_env = ncol(env_current)
  
  cat("\n=== Genomic Offset Calculation ===\n")
  cat("Populations:", n_pop, "\n")
  cat("SNPs (total):", n_snp, "\n")
  cat("Environmental variables:", n_env, "\n")
  
  # Apply SNP filter if provided
  if (!is.null(snp_filter)) {
    allele_freq = allele_freq[, snp_filter, drop = FALSE]
    lfmm_effects = lfmm_effects[snp_filter, , drop = FALSE]
    cat("SNPs (after filtering):", sum(snp_filter), "\n")
  }
  
  # Ensure environmental variables are standardized
  # Use same scaling for current and future based on current
  env_means = colMeans(env_current)
  env_sds = apply(env_current, 2, sd)
  
  env_current_scaled = scale(env_current, center = env_means, scale = env_sds)
  env_future_scaled = scale(env_future, center = env_means, scale = env_sds)
  
  cat("\nEnvironmental variables standardized using current climate statistics\n")
  
  # ========================================================================
  # STEP 1: Calculate genetic values under current climate
  # ========================================================================
  # For each population, calculate adaptive genetic value
  # genetic_value = sum over SNPs and env vars of:
  #   (allele_freq × effect_size × env_value)
  
  genetic_values_current = matrix(0, nrow = n_pop, ncol = n_env)
  
  for (i in 1:n_pop) {
    for (j in 1:n_env) {
      # Allele frequency × effect size for this env variable
      weighted_freqs = allele_freq[i, ] * lfmm_effects[, j]
      # Weighted by environmental value
      genetic_values_current[i, j] = sum(weighted_freqs) * env_current_scaled[i, j]
    }
  }
  
  # Total genetic value across all environmental axes
  genetic_value_current_total = rowSums(genetic_values_current)
  
  cat("\nCurrent genetic values calculated\n")
  
  # ========================================================================
  # STEP 2: Calculate genetic values under future climate
  # ========================================================================
  
  genetic_values_future = matrix(0, nrow = n_pop, ncol = n_env)
  
  for (i in 1:n_pop) {
    for (j in 1:n_env) {
      # Same allele frequencies (no evolution assumed)
      # But different environmental values
      weighted_freqs = allele_freq[i, ] * lfmm_effects[, j]
      genetic_values_future[i, j] = sum(weighted_freqs) * env_future_scaled[i, j]
    }
  }
  
  genetic_value_future_total = rowSums(genetic_values_future)
  
  cat("Future genetic values calculated\n")
  
  # ========================================================================
  # STEP 3: Calculate offset
  # ========================================================================
  
  if (method == "euclidean") {
    # Euclidean distance across all environmental dimensions
    offset = sqrt(rowSums((genetic_values_future - genetic_values_current)^2))
    cat("\nOffset method: Euclidean distance\n")
    
  } else if (method == "absolute") {
    # Simple absolute difference
    offset = abs(genetic_value_future_total - genetic_value_current_total)
    cat("\nOffset method: Absolute difference\n")
    
  } else if (method == "relative") {
    # Relative change
    offset = abs((genetic_value_future_total - genetic_value_current_total) / 
                    (abs(genetic_value_current_total) + 1e-10))
    cat("\nOffset method: Relative change\n")
  }
  
  # ========================================================================
  # STEP 4: Per-variable contributions
  # ========================================================================
  
  # Calculate offset contribution from each environmental variable
  offset_by_var = sqrt((genetic_values_future - genetic_values_current)^2)
  colnames(offset_by_var) = colnames(env_current)
  
  cat("\nOffset calculation complete!\n")
  cat("Mean offset:", round(mean(offset), 4), "\n")
  cat("Range:", round(min(offset), 4), "-", round(max(offset), 4), "\n")
  
  # Return results
  return(list(
    offset = offset,
    offset_by_variable = offset_by_var,
    genetic_values_current = genetic_value_current_total,
    genetic_values_future = genetic_value_future_total,
    genetic_values_current_by_var = genetic_values_current,
    genetic_values_future_by_var = genetic_values_future,
    env_change = env_future_scaled - env_current_scaled,
    n_snps_used = ncol(allele_freq)
  ))
}
```

Calculate the offset now


```{r}
allele_freq = Y_2023
env_current = as.matrix(climate_data_sim_historic[,-1])
rownames(env_current) = pops
env_future = as.matrix(climate_data_sim_future[,-1])
rownames(env_future) = pops
lfmm_effects = mod.lfmm$B
# lfmm_pvalues = pv_2023$pvalue
pop_names = rownames(allele_freq)


offset_all = calculate_genomic_offset(
  allele_freq = allele_freq,
  lfmm_effects = lfmm_effects,
  env_current = env_current,
  env_future = env_future,
  snp_filter = NULL,  # Use all SNPs
  method = "euclidean"
)
```

Plot results (basic)

```{r}
results_df <- data.frame(
  population = pop_names,
  # Offsets
  offset_all_snps = offset_all$offset,
  # Genetic values
  genetic_value_current_all = offset_all$genetic_values_current,
  genetic_value_future_all = offset_all$genetic_values_future,
  # Add environmental change magnitude
  env_change_magnitude = sqrt(rowSums((env_future - env_current)^2))
)
```

```{r}
p1 <- ggplot(results_df, aes(x = env_change_magnitude, y = offset_all_snps, label = population)) +
  geom_point(size = 3, alpha = 0.7, color = "darkred") +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  geom_text_repel()+
  theme_minimal() +
  theme(
    axis.text = element_text(size = 11),
    axis.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    x = "Environmental Change Magnitude",
    y = "Genomic Offset (ALL SNPs)",
    title = "Offset vs Environmental Change"
  )
print(p1)
```

## Make a map of the populations

```{r}
library(maps) # Required to download and plot the map
library(tidyverse)
library(ggthemes)
library(ggdoctheme)
library(ggrepel)

ca_map = map_data("state") |>
  filter(region == "california")

# Read the coordinates and the metadata of the populations
stto_data = read_tsv("data/stto_metadata.tsv")

# Define the population Site_codes to retain
pops = c("BH", "CC", "CP2", "CP3", "DPR", "FR", "IH", "LV1", "LV3", 
"LVTR1", "SC", "SQ1", "SQ2", "SQ3", "TM2", "WL1", "WL2", "WR", 
"YO4", "YO7", "YO8", "YO11")

# create the plotdata dataframe

plotdata = data.frame(Site_code = pops) |>
left_join(stto_data) |>
dplyr::select(Site_code, Lat,	Long,	Elevation, Label)


ggplot() +
  # Plot California base map
  geom_polygon(data = ca_map, 
               aes(x = long, y = lat, group = group),
               fill = "white", color = "black") +
  # Add points for locations
  geom_point(data = plotdata,
             aes(x = Long, y = Lat),
             alpha = 0.7, size = 3, position = position_dodge(width = 1)) +
  # Add labels for locations
  geom_text_repel(data = plotdata,
            aes(x = Long, y = Lat, label = Site_code),
            size = 6) +
  # Customize the appearance
  # Set the coordinate system and zoom
  coord_fixed(1.3) +
  # Customize the theme
  theme_void()

```