---
title: "STTO_climate_model_graphs_2"
output: html_document
date: "2026-05-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load libraries
```{r}
library(tidyverse)
library(ggthemes)
library(ggrepel)
library(patchwork)
library(lfmm)
source("manhattan_plot.R")
library(ggplot2)
```



```{r}
# We would be using this new data and not the old one.
geno_23 = read.delim("data/snp_allele_freq_mat.csv", header = TRUE, sep = ",")
pops =colnames(geno_23)[-1]
npop = length(pops)
```


Read historical data
```{r}
his_data = read.delim("data/Historical_climate_caladapt_1950_2004.tsv", header = TRUE)
```

Read Future Climate data
```{r}
fut_data = read.delim("data/Future_climate_ppt_caladapt_2040_2070.tsv", header = TRUE)
```


Comparing the range of mean environmental variables across all the populations in each different climate model using the HISTORICAL climate data.
```{r}
facet_graph <- 
  his_data %>% 
  filter(dt >= "1990-12-31",
         cvar %in% c("tasmin", "tasmax", "pr"),
         !id %in% c("UCD_Garden", "WL2_Garden")) %>% 
  group_by(id, cvar, gcm) %>% 
  summarise(mean_val = mean(val),
            cv = sd(val)/mean(val),
            sd = sd(val),
            .groups = "drop") %>% 
  pivot_wider(
    names_from = cvar,
    values_from = c(mean_val, sd, cv))
```



GGPLOTS

Mean Precipitation
```{r}
g1 = 
facet_graph %>% 
  ggplot(aes(x = gcm, y = mean_val_pr))+
  geom_crossbar(aes(ymin = mean_val_pr, ymax = mean_val_pr))+
  geom_errorbar(aes(ymin = mean_val_pr - sd_pr, ymax = mean_val_pr + sd_pr), width = 0.6)+
  facet_wrap(~id)+
  labs(title = "Mean Precipitation by Population - Historical", x = "Climate Model", y = "Mean Precipitation")+
  theme(axis.text.x = element_text(angle = 65, hjust=1))
print(g1)
```
```{r}
ggsave("../Figures/mean_pr_bypop.png", plot = g1, width = 10, height = 6)
```


CV Precipitation
```{r}
g2 = 
facet_graph %>% 
  ggplot(aes(x = gcm, y = cv_pr))+
  geom_crossbar(aes(ymin = cv_pr, ymax = cv_pr))+
  facet_wrap(~id)+
  labs(title = "CV Precipitation by Population - Historical", x = "Climate Model", y = "CV Precipitation")+
  theme(axis.text.x = element_text(angle = 65, hjust=1))
print(g2)
```
```{r}
ggsave("../Figures/cv_pr_bypop.png", plot = g2, width = 10, height = 6)
```

Mean Min Temp
```{r}
g3 = 
facet_graph %>% 
  ggplot(aes(x = gcm, y = mean_val_tasmin))+
  geom_crossbar(aes(ymin = mean_val_tasmin, ymax = mean_val_tasmin))+
  geom_errorbar(aes(ymin = mean_val_tasmin - sd_tasmin, ymax = mean_val_tasmin + sd_tasmin), width = 0.6)+
  facet_wrap(~id)+
  labs(title = "Mean Minimum Temperature by Population - Historical", x = "Climate Model", y = "Mean Min Temp")+
  theme(axis.text.x = element_text(angle = 65, hjust=1))
print(g3)
```
```{r}
ggsave("../Figures/mean_min_temp_bypop.png", plot = g3, width = 10, height = 6)
```

CV Minimum Temperature
```{r}
g4 = 
facet_graph %>% 
  ggplot(aes(x = gcm, y = cv_tasmin))+
  geom_crossbar(aes(ymin = cv_tasmin, ymax = cv_tasmin))+
  facet_wrap(~id)+
  labs(title = "CV Minimum Temperature by Population - Historical", x = "Climate Model", y = "CV Min Temp")+
  theme(axis.text.x = element_text(angle = 65, hjust=1))
print(g4)
```
```{r}
ggsave("../Figures/cv_min_temp_bypop.png", plot = g4, width = 10, height = 6)
```

Mean Max Temp
```{r}
g5 = 
facet_graph %>% 
  ggplot(aes(x = gcm, y = mean_val_tasmax))+
  geom_crossbar(aes(ymin = mean_val_tasmax, ymax = mean_val_tasmax))+
  geom_errorbar(aes(ymin = mean_val_tasmax - sd_tasmax, ymax = mean_val_tasmax + sd_tasmax), width = 0.6)+
  facet_wrap(~id)+
  labs(title = "Mean Maximum Temperature by Population - Historical", x = "Climate Model", y = "Mean Max Temp")+
  theme(axis.text.x = element_text(angle = 65, hjust=1))
print(g5)
```
```{r}
ggsave("../Figures/mean_max_temp_bypop.png", plot = g5, width = 10, height = 6)
```

CV Maximum Temperature
```{r}
g6 = 
facet_graph %>% 
  ggplot(aes(x = gcm, y = cv_tasmax))+
  geom_crossbar(aes(ymin = cv_tasmax, ymax = cv_tasmax))+
  facet_wrap(~id)+
  labs(title = "CV Maximum Temperature by Population - Historical", x = "Climate Model", y = "CV Max Temp")+
  theme(axis.text.x = element_text(angle = 65, hjust=1))
print(g6)
```
```{r}
ggsave("../Figures/cv_max_temp.png", plot = g6, width = 10, height = 6)
```


FUTURE Climate Data
```{r}
facet_graph_fut <- 
  fut_data %>% 
  filter(scenario == "rcp85",
         cvar %in% c("tasmin", "tasmax", "pr"),
         !id %in% c("UCD_Garden", "WL2_Garden")) %>% 
  group_by(id, cvar, gcm, scenario) %>% 
  summarise(mean_val = mean(val),
            cv = sd(val)/mean(val),
            sd = sd(val),
            .groups = "drop") %>% 
  pivot_wider(
    names_from = cvar,
    values_from = c(mean_val, sd, cv))
```

Mean Precipitation
```{r}
g7 = 
facet_graph_fut %>% 
  ggplot(aes(x = gcm, y = mean_val_pr))+
  geom_crossbar(aes(ymin = mean_val_pr, ymax = mean_val_pr))+
  geom_errorbar(aes(ymin = mean_val_pr - sd_pr, ymax = mean_val_pr + sd_pr), width = 0.6)+
  facet_wrap(~id)+
  labs(title = "Mean Precipitation by Population - Future(rcp85)", x = "Climate Model", y = "Mean Precipitation")+
  theme(axis.text.x = element_text(angle = 65, hjust=1))
print(g7)
```

```{r}
ggsave("../Figures/fut_mean_pr_bypop.png", plot = g7, width = 10, height = 6)
```


CV Precipitation
```{r}
g8 = 
facet_graph_fut %>% 
  ggplot(aes(x = gcm, y = cv_pr))+
  geom_crossbar(aes(ymin = cv_pr, ymax = cv_pr))+
  facet_wrap(~id)+
  labs(title = "CV Precipitation by Population - Future(rcp85)", x = "Climate Model", y = "CV Precipitation")+
  theme(axis.text.x = element_text(angle = 65, hjust=1))
print(g8)
```
```{r}
ggsave("../Figures/fut_cv_pr_bypop.png", plot = g8, width = 10, height = 6)
```


Mean Minimum Temperature
```{r}
p9 = 
facet_graph_fut %>% 
  ggplot(aes(x = gcm, y = mean_val_tasmin))+
  geom_crossbar(aes(ymin = mean_val_tasmin, ymax = mean_val_tasmin))+
  geom_errorbar(aes(ymin = mean_val_tasmin - sd_tasmin, ymax = mean_val_tasmin + sd_tasmin), width = 0.6)+
  facet_wrap(~id)+
  labs(title = "Mean Minimum Temperature by Population - Future(rcp85)", x = "Climate Model", y = "Mean Min Temp")+
  theme(axis.text.x = element_text(angle = 65, hjust=1))
print(p9)
```
```{r}
ggsave("../Figures/fut_mean_min_temp_bypop.png", plot = p9, width = 10, height = 6)
```


CV Minimum Temperature
```{r}
p10 = 
facet_graph_fut %>% 
  ggplot(aes(x = gcm, y = cv_tasmin))+
  geom_crossbar(aes(ymin = cv_tasmin, ymax = cv_tasmin))+
  facet_wrap(~id)+
  labs(title = "CV Minimum Temperature by Population - Future(rcp85)", x = "Climate Model", y = "CV Min Temp")+
  theme(axis.text.x = element_text(angle = 65, hjust=1))
print(p10)
```
```{r}
ggsave("../Figures/fut_cv_min_temp_bypop.png", plot = p10, width = 10, height = 6)
```

Mean Max Temp
```{r}
p11 = 
facet_graph_fut %>% 
  ggplot(aes(x = gcm, y = mean_val_tasmax))+
  geom_crossbar(aes(ymin = mean_val_tasmax, ymax = mean_val_tasmax))+
  geom_errorbar(aes(ymin = mean_val_tasmax - sd_tasmax, ymax = mean_val_tasmax + sd_tasmax), width = 0.6)+
  facet_wrap(~id)+
  labs(title = "Mean Maximum Temperature by Population - Future(rcp85)", x = "Climate Model", y = "Mean Max Temp")+
  theme(axis.text.x = element_text(angle = 65, hjust=1))
print(p11)
```
```{r}
ggsave("../Figures/fut_mean_max_temp_bypop.png", plot = p11, width = 10, height = 6)
```

CV Maximum Temperature
```{r}
p12 = 
facet_graph_fut %>% 
  ggplot(aes(x = gcm, y = cv_tasmax))+
  geom_crossbar(aes(ymin = cv_tasmax, ymax = cv_tasmax))+
  facet_wrap(~id)+
  labs(title = "CV Maximum Temperature by Population - Future(rcp85)", x = "Climate Model", y = "CV Max Temp")+
  theme(axis.text.x = element_text(angle = 65, hjust=1))
print(p12)
```
```{r}
ggsave("../Figures/fut_cv_max_temp_bypop.png", plot = p12, width = 10, height = 6)
```





