---
title: "STTO_Sim_Part1"
output: html_document
date: "2026-04-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load libraries
```{r}
library(tidyverse)
library(ggthemes)
library(ggrepel)
library(patchwork)
library(lfmm)
source("manhattan_plot.R")
```

```{r}
geno_23 = read.delim("data/merged_maf_common_WL2_2023.tsv", header = TRUE)

# I'm going to simulate some climate data here. But ideally they should be real data
pops =colnames(geno_23)[-1]
npop = length(pops)
climate_data_sim = tibble(PopID = pops,
max_temp = rnorm(npop),
min_temp = rnorm(npop),
cwd = rnorm(npop),
ppt = rnorm(npop))
```


Read historical data
```{r}
his_data = read.delim("data/Historical_climate_caladapt_1950_2004.tsv", header = TRUE)
```

Read Future Climate data
```{r}
fut_data = read.delim("data/Future_climate_ppt_caladapt_2040_2070.tsv", header = TRUE)
```

```{r}
mean_fut <- fut_data %>% 
  group_by(id, cvar, gcm) %>% 
  summarise(mean_val = mean(val))
```
```{r}
write_csv(mean_fut, "data/mean_future_climate.csv")
```


```{r}
mean_his <- his_data %>% 
  group_by(id, cvar, gcm) %>% 
  summarise(mean_val = mean(val))
```
```{r}
write_csv(mean_his, "data/mean_historical_climate.csv")
```




HadGEM2-ES Model
```{r}
HadGEM2_mean = mean_his %>% 
  filter(gcm == "HadGEM2-ES") %>% 
  pivot_wider(names_from = cvar, values_from = mean_val)
```

do this once
```{r}
Y_2023 = as.matrix(geno_23[,-1])
Y_2023 = t(Y_2023)
```

do this everytime running a model
```{r}
X_2023 = HadGEM2_mean %>% 
  ungroup() %>% 
  select(tasmax, tasmin, pr) %>% 
  as.matrix() %>% 
  scale()
```


```{r}
rownames(X_2023) = HadGEM2_mean$id
common = intersect(rownames(X_2023), rownames(Y_2023))
Y_2023 = Y_2023[common,]
X_2023 = X_2023[common, , drop = FALSE]
```

```{r}
mod.lfmm = lfmm_ridge(Y = Y_2023, 
    X = X_2023, 
    K = 1)

pv_2023 = lfmm_test(Y = Y_2023, 
    X = X_2023, 
    lfmm = mod.lfmm, 
    calibrate = "gif")
```


```{r}
pvalues_2023.raw = pv_2023$pvalue[,1]

n = length(sort(pvalues_2023.raw))
df = tibble(observed = sort(-log10(pvalues_2023.raw)),
expected = sort(-log10(ppoints(n))),
clower   = -log10(qbeta(p = (1 - 0.95) / 2, shape1 = 1:n, shape2 = n:1)),
cupper   = -log10(qbeta(p = (1 + 0.95) / 2, shape1 = 1:n, shape2 = n:1))
)

# Convert p-values to 1-df chi-squared statistics
chi2 <- qchisq(pvalues_2023.raw, df=1, lower.tail=FALSE)
# Calculate lambda
lambda <- median(chi2) / qchisq(0.5, df=1)
print(lambda)


df |>
ggplot() +
geom_ribbon(aes(x = expected, ymin = clower, ymax = cupper), alpha = 0.1) +
geom_point(aes(expected, observed), shape = 1, size = 3) +
geom_abline(intercept = 0, slope = 1, alpha = 0.5) +
theme_bw() +
labs(x = expression(paste("Expected -log"[10], plain(P))),
y = expression(paste("Observed -log"[10], plain(P))))
```

```{r}
 mod.lfmm_lasso <- lfmm_lasso(Y = Y_2023, 
                        X = X_2023, 
                        K = 5,
                        nozero.prop = 0.01)
```


```{r}
pv_lasso <- lfmm_test(Y = Y_2023, 
                 X = X_2023, 
                 lfmm = mod.lfmm_lasso, 
                 calibrate = "gif")
```

p-values for max temp/min temp/pr
```{r}
pvalues_2023.raw_lasso = pv_lasso$pvalue[,3]

n = length(sort(pvalues_2023.raw_lasso))
df = tibble(observed = sort(-log10(pvalues_2023.raw_lasso)),
expected = sort(-log10(ppoints(n))),
clower   = -log10(qbeta(p = (1 - 0.95) / 2, shape1 = 1:n, shape2 = n:1)),
cupper   = -log10(qbeta(p = (1 + 0.95) / 2, shape1 = 1:n, shape2 = n:1))
)

# Convert p-values to 1-df chi-squared statistics
chi2 <- qchisq(pvalues_2023.raw_lasso, df=1, lower.tail=FALSE)
# Calculate lambda
lambda <- median(chi2) / qchisq(0.5, df=1)
print(lambda)


df |>
ggplot() +
geom_ribbon(aes(x = expected, ymin = clower, ymax = cupper), alpha = 0.1) +
geom_point(aes(expected, observed), shape = 1, size = 3) +
geom_abline(intercept = 0, slope = 1, alpha = 0.5) +
theme_bw() +
labs(x = expression(paste("Expected -log"[10], plain(P))),
y = expression(paste("Observed -log"[10], plain(P))))
```

Function
```{r}
plot_qq <- function(pvalue_data = NULL,
                    index = 1){
  pvalues.raw = pvalue_data$pvalue[,index]
  
n = length(sort(pvalues.raw))
df = tibble(observed = sort(-log10(pvalues.raw)),
expected = sort(-log10(ppoints(n))),
clower   = -log10(qbeta(p = (1 - 0.95) / 2, shape1 = 1:n, shape2 = n:1)),
cupper   = -log10(qbeta(p = (1 + 0.95) / 2, shape1 = 1:n, shape2 = n:1))
)

# Convert p-values to 1-df chi-squared statistics
chi2 <- qchisq(pvalues.raw, df=1, lower.tail=FALSE)
# Calculate lambda
lambda <- median(chi2) / qchisq(0.5, df=1)
print(lambda)

y_max_print = df |>
  slice_max(observed, n=1) |>
  pull(observed)
  lambda = round(lambda, digits =2)


p = df |>
ggplot() +
geom_ribbon(aes(x = expected, ymin = clower, ymax = cupper), alpha = 0.1) +
geom_point(aes(expected, observed), shape = 1, size = 3) +
geom_abline(intercept = 0, slope = 1, alpha = 0.5) +
theme_bw() +
annotate(geom = "text", x=1, y = y_max_print-0.5, label = paste0("Lambda = ", lambda))+
labs(x = expression(paste("Expected -log"[10], plain(P))),
y = expression(paste("Observed -log"[10], plain(P))))

return(p)

}
```

```{r}
plot_qq(pv_lasso, index = 1)
```


```{r}
run_lasso <- function(X=NULL,Y=NULL,K=4) {
   mod.lfmm_lasso <- lfmm_lasso(Y = Y, 
                        X = X, 
                        K = K,
                        nozero.prop = 0.01)
   pv_lasso <- lfmm_test(Y = Y, 
                 X = X, 
                 lfmm = mod.lfmm_lasso, 
                 calibrate = "gif")
   return(pv_lasso)
}
```

```{r}
pv_lasso_1 = run_lasso(X_2023, Y_2023)
```

```{r}
plot_qq(pv_lasso_1, index = 1)
plot_qq(pv_lasso_1, index = 2)
plot_qq(pv_lasso_1, index = 3)
```


```{r}
plot_data_2023 = tibble(loci = geno_23[,1], p_val = pv_lasso_1$pvalue[,1], calibrated.pvalue = pv_lasso_1$calibrated.pvalue[,1]) |>
separate_wider_delim(loci, delim = ":", names = c("chr", "bp")) |>
filter(str_detect(chr, "Chr")) |>
mutate(p.bonf = p.adjust(p_val, method = "bonferroni"),
bp = as.numeric(bp))


manhattan_WL2 = plot_manhattan(plot_data_2023,
                        chr_col = "chr",
                        pos_col = "bp",
                        value_col = "calibrated.pvalue",
                        transform = "neglog10",
                        title = "WL2 2023",
                        highlight_top_n = 20)

print(manhattan_WL2)

# Optionally save plot
#ggsave("WL2_2023.png", manhattan_WL2, width = 12, height = 6, dpi = 300)
```


CNRM_CM5 Model
```{r}
CNRM_CM5_mean = mean_his %>% 
  filter(gcm == "CNRM-CM5") %>% 
  pivot_wider(names_from = cvar, values_from = mean_val)
```


```{r}
X_CNRM_CM5 = CNRM_CM5_mean %>% 
  ungroup() %>% 
  select(tasmax, tasmin, pr) %>% 
  as.matrix() %>% 
  scale()
```

```{r}
rownames(X_CNRM_CM5) = CNRM_CM5_mean$id
common = intersect(rownames(X_CNRM_CM5), rownames(Y_2023))
Y_2023 = Y_2023[common,]
X_CNRM_CM5 = X_CNRM_CM5[common, , drop = FALSE]
```

```{r}
pv_lasso_CNRM_CM5 = run_lasso(X_CNRM_CM5, Y_2023)
```

```{r}
plot_qq(pv_lasso_CNRM_CM5, index = 1)
plot_qq(pv_lasso_CNRM_CM5, index = 2)
plot_qq(pv_lasso_CNRM_CM5, index = 3)
```

```{r}
plot_data_2023 = tibble(loci = geno_23[,1], p_val = pv_lasso_CNRM_CM5$pvalue[,3], calibrated.pvalue = pv_lasso_CNRM_CM5$calibrated.pvalue[,3]) |>
separate_wider_delim(loci, delim = ":", names = c("chr", "bp")) |>
filter(str_detect(chr, "Chr")) |>
mutate(p.bonf = p.adjust(p_val, method = "bonferroni"),
bp = as.numeric(bp))


manhattan_WL2 = plot_manhattan(plot_data_2023,
                        chr_col = "chr",
                        pos_col = "bp",
                        value_col = "calibrated.pvalue",
                        transform = "neglog10",
                        title = "WL2 2023",
                        highlight_top_n = 20)

print(manhattan_WL2)

# Optionally save plot
#ggsave("WL2_2023.png", manhattan_WL2, width = 12, height = 6, dpi = 300)
```


CanESM2 Model
```{r}
CanESM2_mean = mean_his %>% 
  filter(gcm == "CanESM2") %>% 
  pivot_wider(names_from = cvar, values_from = mean_val)
```


```{r}
X_CanESM2 = CanESM2_mean %>% 
  ungroup() %>% 
  select(tasmax, tasmin, pr) %>% 
  as.matrix() %>% 
  scale()
```

```{r}
rownames(X_CanESM2) = CanESM2_mean$id
common = intersect(rownames(X_CanESM2), rownames(Y_2023))
Y_2023 = Y_2023[common,]
X_CanESM2 = X_CanESM2[common, , drop = FALSE]
```

```{r}
pv_lasso_CanESM2 = run_lasso(X_CanESM2, Y_2023)
```

```{r}
plot_qq(pv_lasso_CanESM2, index = 1)
plot_qq(pv_lasso_CanESM2, index = 2)
plot_qq(pv_lasso_CanESM2, index = 3)
```

```{r}
plot_data_2023 = tibble(loci = geno_23[,1], p_val = pv_lasso_CanESM2$pvalue[,3], calibrated.pvalue = pv_lasso_CanESM2$calibrated.pvalue[,3]) |>
separate_wider_delim(loci, delim = ":", names = c("chr", "bp")) |>
filter(str_detect(chr, "Chr")) |>
mutate(p.bonf = p.adjust(p_val, method = "bonferroni"),
bp = as.numeric(bp))


manhattan_WL2 = plot_manhattan(plot_data_2023,
                        chr_col = "chr",
                        pos_col = "bp",
                        value_col = "calibrated.pvalue",
                        transform = "neglog10",
                        title = "WL2 2023",
                        highlight_top_n = 20)

print(manhattan_WL2)

# Optionally save plot
#ggsave("WL2_2023.png", manhattan_WL2, width = 12, height = 6, dpi = 300)
```


MIROC5 Model

```{r}
MIROC5_mean = mean_his %>% 
  filter(gcm == "MIROC5") %>% 
  pivot_wider(names_from = cvar, values_from = mean_val)
```

```{r}
MIROC5_fut_mean = mean_fut %>% 
  filter(gcm == "MIROC5") %>% 
  pivot_wider(names_from = cvar, values_from = mean_val)
```


```{r}
X_MIROC5 = MIROC5_mean %>% 
  ungroup() %>% 
  select(tasmax, tasmin, pr) %>% 
  as.matrix() %>% 
  scale()
```

```{r}
X_MIROC5_fut = MIROC5_fut_mean %>% 
  ungroup() %>% 
  select(tasmax, tasmin, pr) %>% 
  as.matrix() %>% 
  scale()
```


```{r}
rownames(X_MIROC5) = MIROC5_mean$id
common = intersect(rownames(X_MIROC5), rownames(Y_2023))
Y_2023 = Y_2023[common,]
X_MIROC5 = X_MIROC5[common, , drop = FALSE]
```

```{r}
rownames(X_MIROC5_fut) = MIROC5_fut_mean$id
X_MIROC5_fut = X_MIROC5_fut[common, , drop = FALSE]
```


```{r}
pv_lasso_MIROC5 = run_lasso(X_MIROC5, Y_2023)
```


```{r}
plot_qq(pv_lasso_MIROC5, index = 1)
plot_qq(pv_lasso_MIROC5, index = 2)
plot_qq(pv_lasso_MIROC5, index = 3)
```


```{r}
plot_data_2023 = tibble(loci = geno_23[,1], p_val = pv_lasso_MIROC5$pvalue[,3], calibrated.pvalue = pv_lasso_MIROC5$calibrated.pvalue[,3]) |>
separate_wider_delim(loci, delim = ":", names = c("chr", "bp")) |>
filter(str_detect(chr, "Chr")) |>
mutate(p.bonf = p.adjust(p_val, method = "bonferroni"),
bp = as.numeric(bp))


manhattan_WL2 = plot_manhattan(plot_data_2023,
                        chr_col = "chr",
                        pos_col = "bp",
                        value_col = "calibrated.pvalue",
                        transform = "neglog10",
                        title = "WL2 2023",
                        highlight_top_n = 20)

print(manhattan_WL2)

# Optionally save plot
#ggsave("WL2_2023.png", manhattan_WL2, width = 12, height = 6, dpi = 300)
```


Genetic Offset

MIROC5 Model
```{r}
calculate_genomic_offset = function(allele_freq, 
                                     lfmm_effects, 
                                     env_current, 
                                     env_future,
                                     snp_filter = NULL,
                                     method = "euclidean") {
  
  # Dimensions check
  n_pop = nrow(allele_freq)
  n_snp = ncol(allele_freq)
  n_env = ncol(env_current)
  
  cat("\n=== Genomic Offset Calculation ===\n")
  cat("Populations:", n_pop, "\n")
  cat("SNPs (total):", n_snp, "\n")
  cat("Environmental variables:", n_env, "\n")
  
  # Apply SNP filter if provided
  if (!is.null(snp_filter)) {
    allele_freq = allele_freq[, snp_filter, drop = FALSE]
    lfmm_effects = lfmm_effects[snp_filter, , drop = FALSE]
    cat("SNPs (after filtering):", sum(snp_filter), "\n")
  }
  
  # Ensure environmental variables are standardized
  # Use same scaling for current and future based on current
  env_means = colMeans(env_current)
  env_sds = apply(env_current, 2, sd)
  
  env_current_scaled = scale(env_current, center = env_means, scale = env_sds)
  env_future_scaled = scale(env_future, center = env_means, scale = env_sds)
  
  cat("\nEnvironmental variables standardized using current climate statistics\n")
  
  # ========================================================================
  # STEP 1: Calculate genetic values under current climate
  # ========================================================================
  # For each population, calculate adaptive genetic value
  # genetic_value = sum over SNPs and env vars of:
  #   (allele_freq × effect_size × env_value)
  
  genetic_values_current = matrix(0, nrow = n_pop, ncol = n_env)
  
  for (i in 1:n_pop) {
    for (j in 1:n_env) {
      # Allele frequency × effect size for this env variable
      weighted_freqs = allele_freq[i, ] * lfmm_effects[, j]
      # Weighted by environmental value
      genetic_values_current[i, j] = sum(weighted_freqs) * env_current_scaled[i, j]
    }
  }
  
  # Total genetic value across all environmental axes
  genetic_value_current_total = rowSums(genetic_values_current)
  
  cat("\nCurrent genetic values calculated\n")
  
  # ========================================================================
  # STEP 2: Calculate genetic values under future climate
  # ========================================================================
  
  genetic_values_future = matrix(0, nrow = n_pop, ncol = n_env)
  
  for (i in 1:n_pop) {
    for (j in 1:n_env) {
      # Same allele frequencies (no evolution assumed)
      # But different environmental values
      weighted_freqs = allele_freq[i, ] * lfmm_effects[, j]
      genetic_values_future[i, j] = sum(weighted_freqs) * env_future_scaled[i, j]
    }
  }
  
  genetic_value_future_total = rowSums(genetic_values_future)
  
  cat("Future genetic values calculated\n")
  
  # ========================================================================
  # STEP 3: Calculate offset
  # ========================================================================
  
  if (method == "euclidean") {
    # Euclidean distance across all environmental dimensions
    offset = sqrt(rowSums((genetic_values_future - genetic_values_current)^2))
    cat("\nOffset method: Euclidean distance\n")
    
  } else if (method == "absolute") {
    # Simple absolute difference
    offset = abs(genetic_value_future_total - genetic_value_current_total)
    cat("\nOffset method: Absolute difference\n")
    
  } else if (method == "relative") {
    # Relative change
    offset = abs((genetic_value_future_total - genetic_value_current_total) / 
                    (abs(genetic_value_current_total) + 1e-10))
    cat("\nOffset method: Relative change\n")
  }
  
  # ========================================================================
  # STEP 4: Per-variable contributions
  # ========================================================================
  
  # Calculate offset contribution from each environmental variable
  offset_by_var = sqrt((genetic_values_future - genetic_values_current)^2)
  colnames(offset_by_var) = colnames(env_current)
  
  cat("\nOffset calculation complete!\n")
  cat("Mean offset:", round(mean(offset), 4), "\n")
  cat("Range:", round(min(offset), 4), "-", round(max(offset), 4), "\n")
  
  # Return results
  return(list(
    offset = offset,
    offset_by_variable = offset_by_var,
    genetic_values_current = genetic_value_current_total,
    genetic_values_future = genetic_value_future_total,
    genetic_values_current_by_var = genetic_values_current,
    genetic_values_future_by_var = genetic_values_future,
    env_change = env_future_scaled - env_current_scaled,
    n_snps_used = ncol(allele_freq)
  ))
}
```



```{r}
allele_freq = Y_2023
env_current = as.matrix(X_MIROC5)
rownames(env_current) = pops
env_future = as.matrix(X_MIROC5_fut)
rownames(env_future) = pops
lfmm_effects = pv_lasso_MIROC5$B
# lfmm_pvalues = pv_2023$pvalue
pop_names = rownames(allele_freq)


offset_all = calculate_genomic_offset(
  allele_freq = allele_freq,
  lfmm_effects = lfmm_effects,
  env_current = env_current,
  env_future = env_future,
  snp_filter = NULL,  # Use all SNPs
  method = "euclidean"
)
```

```{r}
results_df <- data.frame(
  population = pop_names,
  # Offsets
  offset_all_snps = offset_all$offset,
  # Genetic values
  genetic_value_current_all = offset_all$genetic_values_current,
  genetic_value_future_all = offset_all$genetic_values_future,
  # Add environmental change magnitude
  env_change_magnitude = sqrt(rowSums((env_future - env_current)^2))
)
```

```{r}
p1 <- ggplot(results_df, aes(x = env_change_magnitude, y = offset_all_snps, label = population)) +
  geom_point(size = 3, alpha = 0.7, color = "darkred") +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  geom_text_repel()+
  theme_minimal() +
  theme(
    axis.text = element_text(size = 11),
    axis.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    x = "Environmental Change Magnitude",
    y = "Genomic Offset (ALL SNPs)",
    title = "Offset vs Environmental Change"
  )
print(p1)
```

CNRM_CM5 Model

```{r}
CNRM_CM5_mean = mean_his %>% 
  filter(gcm == "CNRM-CM5") %>% 
  pivot_wider(names_from = cvar, values_from = mean_val)
```

```{r}
CNRM_CM5_fut_mean = mean_fut %>% 
  filter(gcm == "CNRM-CM5") %>% 
  pivot_wider(names_from = cvar, values_from = mean_val)
```


```{r}
X_CNRM_CM5 = CNRM_CM5_mean %>% 
  ungroup() %>% 
  select(tasmax, tasmin, pr) %>% 
  as.matrix() %>% 
  scale()
```

```{r}
X_CNRM_CM5_fut = CNRM_CM5_fut_mean %>% 
  ungroup() %>% 
  select(tasmax, tasmin, pr) %>% 
  as.matrix() %>% 
  scale()
```


```{r}
rownames(X_CNRM_CM5) = CNRM_CM5_mean$id
common = intersect(rownames(X_CNRM_CM5), rownames(Y_2023))
Y_2023 = Y_2023[common,]
X_CNRM_CM5 = X_CNRM_CM5[common, , drop = FALSE]
```

```{r}
rownames(X_CNRM_CM5_fut) = CNRM_CM5_fut_mean$id
X_CNRM_CM5_fut = X_CNRM_CM5_fut[common, , drop = FALSE]
```


```{r}
allele_freq = Y_2023
env_current = as.matrix(X_CNRM_CM5)
rownames(env_current) = pops
env_future = as.matrix(X_CNRM_CM5_fut)
rownames(env_future) = pops
lfmm_effects = pv_lasso_CNRM_CM5$B
# lfmm_pvalues = pv_2023$pvalue
pop_names = rownames(allele_freq)


offset_all = calculate_genomic_offset(
  allele_freq = allele_freq,
  lfmm_effects = lfmm_effects,
  env_current = env_current,
  env_future = env_future,
  snp_filter = NULL,  # Use all SNPs
  method = "euclidean"
)
```
```{r}
results_df <- data.frame(
  population = pop_names,
  # Offsets
  offset_all_snps = offset_all$offset,
  # Genetic values
  genetic_value_current_all = offset_all$genetic_values_current,
  genetic_value_future_all = offset_all$genetic_values_future,
  # Add environmental change magnitude
  env_change_magnitude = sqrt(rowSums((env_future - env_current)^2))
)
```

```{r}
p1 <- ggplot(results_df, aes(x = env_change_magnitude, y = offset_all_snps, label = population)) +
  geom_point(size = 3, alpha = 0.7, color = "darkred") +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  geom_text_repel()+
  theme_minimal() +
  theme(
    axis.text = element_text(size = 11),
    axis.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    x = "Environmental Change Magnitude",
    y = "Genomic Offset (ALL SNPs)",
    title = "Offset vs Environmental Change"
  )
print(p1)
```

CanESM2 Model


```{r}
CanESM2_mean = mean_his %>% 
  filter(gcm == "CanESM2") %>% 
  pivot_wider(names_from = cvar, values_from = mean_val)
```

```{r}
CanESM2_fut_mean = mean_fut %>% 
  filter(gcm == "CanESM2") %>% 
  pivot_wider(names_from = cvar, values_from = mean_val)
```


```{r}
X_CanESM2 = CanESM2_mean %>% 
  ungroup() %>% 
  select(tasmax, tasmin, pr) %>% 
  as.matrix() %>% 
  scale()
```

```{r}
X_CanESM2_fut = CanESM2_fut_mean %>% 
  ungroup() %>% 
  select(tasmax, tasmin, pr) %>% 
  as.matrix() %>% 
  scale()
```


```{r}
rownames(X_CanESM2) = CanESM2_mean$id
common = intersect(rownames(X_CNRM_CM5), rownames(Y_2023))
Y_2023 = Y_2023[common,]
X_CanESM2 = X_CanESM2[common, , drop = FALSE]
```

```{r}
rownames(X_CanESM2_fut) = CanESM2_fut_mean$id
X_CanESM2_fut = X_CanESM2_fut[common, , drop = FALSE]
```


```{r}
allele_freq = Y_2023
env_current = as.matrix(X_CanESM2)
rownames(env_current) = pops
env_future = as.matrix(X_CanESM2_fut)
rownames(env_future) = pops
lfmm_effects = pv_lasso_CanESM2$B
# lfmm_pvalues = pv_2023$pvalue
pop_names = rownames(allele_freq)


offset_all = calculate_genomic_offset(
  allele_freq = allele_freq,
  lfmm_effects = lfmm_effects,
  env_current = env_current,
  env_future = env_future,
  snp_filter = NULL,  # Use all SNPs
  method = "euclidean"
)
```

```{r}
results_df <- data.frame(
  population = pop_names,
  # Offsets
  offset_all_snps = offset_all$offset,
  # Genetic values
  genetic_value_current_all = offset_all$genetic_values_current,
  genetic_value_future_all = offset_all$genetic_values_future,
  # Add environmental change magnitude
  env_change_magnitude = sqrt(rowSums((env_future - env_current)^2))
)
```

```{r}
p1 <- ggplot(results_df, aes(x = env_change_magnitude, y = offset_all_snps, label = population)) +
  geom_point(size = 3, alpha = 0.7, color = "darkred") +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  geom_text_repel()+
  theme_minimal() +
  theme(
    axis.text = element_text(size = 11),
    axis.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    x = "Environmental Change Magnitude",
    y = "Genomic Offset (ALL SNPs)",
    title = "Offset vs Environmental Change"
  )
print(p1)
```


HadGEM2-ES Model

```{r}
HadGEM2_mean = mean_his %>% 
  filter(gcm == "HadGEM2-ES") %>% 
  pivot_wider(names_from = cvar, values_from = mean_val)
```

```{r}
HadGEM2_fut_mean = mean_fut %>% 
  filter(gcm == "HadGEM2-ES") %>% 
  pivot_wider(names_from = cvar, values_from = mean_val)
```


```{r}
X_HadGEM2 = HadGEM2_mean %>% 
  ungroup() %>% 
  select(tasmax, tasmin, pr) %>% 
  as.matrix() %>% 
  scale()
```

```{r}
X_HadGEM2_fut = HadGEM2_fut_mean %>% 
  ungroup() %>% 
  select(tasmax, tasmin, pr) %>% 
  as.matrix() %>% 
  scale()
```


```{r}
rownames(X_HadGEM2) = HadGEM2_mean$id
common = intersect(rownames(X_HadGEM2), rownames(Y_2023))
Y_2023 = Y_2023[common,]
X_HadGEM2 = X_HadGEM2[common, , drop = FALSE]
```

```{r}
rownames(X_HadGEM2_fut) = HadGEM2_fut_mean$id
X_HadGEM2_fut = X_HadGEM2_fut[common, , drop = FALSE]
```


```{r}
allele_freq = Y_2023
env_current = as.matrix(X_HadGEM2)
rownames(env_current) = pops
env_future = as.matrix(X_HadGEM2_fut)
rownames(env_future) = pops
lfmm_effects = pv_lasso_HadGEM2$B
# lfmm_pvalues = pv_2023$pvalue
pop_names = rownames(allele_freq)


offset_all = calculate_genomic_offset(
  allele_freq = allele_freq,
  lfmm_effects = lfmm_effects,
  env_current = env_current,
  env_future = env_future,
  snp_filter = NULL,  # Use all SNPs
  method = "euclidean"
)
```

```{r}
results_df <- data.frame(
  population = pop_names,
  # Offsets
  offset_all_snps = offset_all$offset,
  # Genetic values
  genetic_value_current_all = offset_all$genetic_values_current,
  genetic_value_future_all = offset_all$genetic_values_future,
  # Add environmental change magnitude
  env_change_magnitude = sqrt(rowSums((env_future - env_current)^2))
)
```

```{r}
p1 <- ggplot(results_df, aes(x = env_change_magnitude, y = offset_all_snps, label = population)) +
  geom_point(size = 3, alpha = 0.7, color = "darkred") +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  geom_text_repel()+
  theme_minimal() +
  theme(
    axis.text = element_text(size = 11),
    axis.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    x = "Environmental Change Magnitude",
    y = "Genomic Offset (ALL SNPs)",
    title = "Offset vs Environmental Change"
  )
print(p1)
```





















SNP_allele data

```{r}
geno_23 = read.delim("data/snp_allele_freq_mat.csv", header = TRUE, sep = "," )
```
CNRM_CM5 Model
```{r}
Y_2023 = as.matrix(geno_23[,-1])
Y_2023 = t(Y_2023)
```


```{r}
CNRM_CM5_mean = mean_his %>% 
  filter(gcm == "CNRM-CM5") %>% 
  pivot_wider(names_from = cvar, values_from = mean_val)
```


```{r}
X_CNRM_CM5 = CNRM_CM5_mean %>% 
  ungroup() %>% 
  select(tasmax, tasmin, pr) %>% 
  as.matrix() %>% 
  scale()
```

```{r}
rownames(X_CNRM_CM5) = CNRM_CM5_mean$id
common = intersect(rownames(X_CNRM_CM5), rownames(Y_2023))
Y_2023 = Y_2023[common,]
X_CNRM_CM5 = X_CNRM_CM5[common, , drop = FALSE]
```

```{r}
pv_lasso_CNRM_CM5 = run_lasso(X_CNRM_CM5, Y_2023)
```

```{r}
plot_qq(pv_lasso_CNRM_CM5, index = 1)
plot_qq(pv_lasso_CNRM_CM5, index = 2)
plot_qq(pv_lasso_CNRM_CM5, index = 3)
```

```{r}
plot_data_2023 = tibble(loci = geno_23[,1], p_val = pv_lasso_CNRM_CM5$pvalue[,3], calibrated.pvalue = pv_lasso_CNRM_CM5$calibrated.pvalue[,3]) |>
separate_wider_delim(loci, delim = "_", names = c("chr", "bp")) |>
filter(str_detect(chr, "Chr")) |>
mutate(p.bonf = p.adjust(p_val, method = "bonferroni"),
bp = as.numeric(bp))


manhattan_WL2 = plot_manhattan(plot_data_2023,
                        chr_col = "chr",
                        pos_col = "bp",
                        value_col = "calibrated.pvalue",
                        transform = "neglog10",
                        title = "WL2 2023",
                        highlight_top_n = 20)

print(manhattan_WL2)

# Optionally save plot
#ggsave("WL2_2023.png", manhattan_WL2, width = 12, height = 6, dpi = 300)
```


CanESM2 MODEL
```{r}
CanESM2_mean = mean_his %>% 
  filter(gcm == "CanESM2") %>% 
  pivot_wider(names_from = cvar, values_from = mean_val)
```


```{r}
X_CanESM2 = CanESM2_mean %>% 
  ungroup() %>% 
  select(tasmax, tasmin, pr) %>% 
  as.matrix() %>% 
  scale()
```

```{r}
rownames(X_CanESM2) = CanESM2_mean$id
common = intersect(rownames(X_CanESM2), rownames(Y_2023))
Y_2023 = Y_2023[common,]
X_CanESM2 = X_CanESM2[common, , drop = FALSE]
```

```{r}
pv_lasso_CanESM2 = run_lasso(X_CanESM2, Y_2023)
```

slightly higher values 
```{r}
plot_qq(pv_lasso_CanESM2, index = 1)
plot_qq(pv_lasso_CanESM2, index = 2)
plot_qq(pv_lasso_CanESM2, index = 3)
```
```{r}
plot_data_2023 = tibble(loci = geno_23[,1], p_val = pv_lasso_CanESM2$pvalue[,3], calibrated.pvalue = pv_lasso_CanESM2$calibrated.pvalue[,3]) |>
separate_wider_delim(loci, delim = "_", names = c("chr", "bp")) |>
filter(str_detect(chr, "Chr")) |>
mutate(p.bonf = p.adjust(p_val, method = "bonferroni"),
bp = as.numeric(bp))


manhattan_WL2 = plot_manhattan(plot_data_2023,
                        chr_col = "chr",
                        pos_col = "bp",
                        value_col = "calibrated.pvalue",
                        transform = "neglog10",
                        title = "WL2 2023",
                        highlight_top_n = 20)

print(manhattan_WL2)

# Optionally save plot
#ggsave("WL2_2023.png", manhattan_WL2, width = 12, height = 6, dpi = 300)
```



MIROC Model
```{r}
MIROC5_mean = mean_his %>% 
  filter(gcm == "MIROC5") %>% 
  pivot_wider(names_from = cvar, values_from = mean_val)
```


```{r}
X_MIROC5 = MIROC5_mean %>% 
  ungroup() %>% 
  select(tasmax, tasmin, pr) %>% 
  as.matrix() %>% 
  scale()
```

```{r}
rownames(X_MIROC5) = MIROC5_mean$id
common = intersect(rownames(X_MIROC5), rownames(Y_2023))
Y_2023 = Y_2023[common,]
X_MIROC5 = X_MIROC5[common, , drop = FALSE]
```

```{r}
pv_lasso_MIROC5 = run_lasso(X_MIROC5, Y_2023)
```


```{r}
plot_qq(pv_lasso_MIROC5, index = 1)
plot_qq(pv_lasso_MIROC5, index = 2)
plot_qq(pv_lasso_MIROC5, index = 3)
```

```{r}
plot_data_2023 = tibble(loci = geno_23[,1], p_val = pv_lasso_MIROC5$pvalue[,3], calibrated.pvalue = pv_lasso_MIROC5$calibrated.pvalue[,3]) |>
separate_wider_delim(loci, delim = "_", names = c("chr", "bp")) |>
filter(str_detect(chr, "Chr")) |>
mutate(p.bonf = p.adjust(p_val, method = "bonferroni"),
bp = as.numeric(bp))


manhattan_WL2 = plot_manhattan(plot_data_2023,
                        chr_col = "chr",
                        pos_col = "bp",
                        value_col = "calibrated.pvalue",
                        transform = "neglog10",
                        title = "WL2 2023",
                        highlight_top_n = 20)

print(manhattan_WL2)

# Optionally save plot
#ggsave("WL2_2023.png", manhattan_WL2, width = 12, height = 6, dpi = 300)
```



HadGEM2-ES Model
```{r}
HadGEM2_mean = mean_his %>% 
  filter(gcm == "HadGEM2-ES") %>% 
  pivot_wider(names_from = cvar, values_from = mean_val)
```


```{r}
X_HadGEM2 = HadGEM2_mean %>% 
  ungroup() %>% 
  select(tasmax, tasmin, pr) %>% 
  as.matrix() %>% 
  scale()
```

```{r}
rownames(X_HadGEM2) = HadGEM2_mean$id
common = intersect(rownames(X_HadGEM2), rownames(Y_2023))
Y_2023 = Y_2023[common,]
X_HadGEM2 = X_HadGEM2[common, , drop = FALSE]
```

```{r}
pv_lasso_HadGEM2 = run_lasso(X_HadGEM2, Y_2023)
```

